\section{Introduction}
\label{sec:intro}

Medical question-answering benchmarks have proliferated, but maternal,
neonatal, child, and reproductive health---and the sub-Saharan, midwife-led
primary-care setting in particular---are a thin slice of them. The widely used
sets draw on US and Indian licensing exams (MedQA~\cite{jin2021medqa},
MedMCQA~\cite{pal2022medmcqa}) or broad clinical chat
(HealthBench~\cite{arora2025healthbench}); AfriMed-QA~\cite{olatunji2025afrimedqa}
is pan-African but multi-specialty. None is scoped to the maternal, neonatal, child,
and reproductive-health questions a nurse-midwife actually asks at the point of
care.

Retrieval is worse served. Medical retrieval-augmented-generation benchmarks
such as MIRAGE~\cite{xiong2024benchmarking} score the \emph{end-to-end} answer
drawn from a corpus, not whether an individual retrieved passage is relevant;
the classical medical-IR relevance collections (the TREC Clinical Decision
Support and Precision Medicine tracks) judge full PubMed articles against case
descriptions, not the short guideline chunks a deployed RAG system retrieves and
injects into a prompt. To our knowledge there is no public chunk-level relevance
benchmark for maternal-health guideline retrieval.

We release two benchmarks%
\footnote{Datasets: \url{https://huggingface.co/datasets/nmrenyi/mamabench},
\url{https://huggingface.co/datasets/nmrenyi/mamaretrieval}. Construction code:
\url{https://github.com/nmrenyi/mamabench},
\url{https://github.com/nmrenyi/mamaretrieval}. Released for research use; some
sources are non-commercial (e.g.\ AfriMed-QA, CC~BY-NC-SA~4.0), with each item's
license recorded in its per-source manifest.}
that close these gaps for the domain. Both were built
to supply the test data for a companion system paper~\cite{ren2026mamai}---a
deployed, fully on-device RAG assistant for nurse-midwives in Zanzibar---but
each is usable as a standalone instrument.
\begin{itemize}
  \item \textbf{mamabench} (\S\ref{sec:mamabench}) is a scope-filtered
  question-answering benchmark: 25{,}949 items assembled from seven
  expert-authored sources across multiple-choice, short-answer, and
  rubric-graded tracks, plus a 6{,}853-triple judge-calibration side-file---not
  labels of our own, but HealthBench's physician annotations re-scoped to this
  domain---for vetting an LLM judge before it is trusted to score the open-ended
  items. Its scope classifier is cross-checked by agreement with a larger model
  and parity against prior labels.
  \item \textbf{mamaretrieval} (\S\ref{sec:mamaretrieval}) pairs 3{,}185 clinical
  queries with graded ($0$--$6$) relevance labels over a 63{,}650-chunk
  maternal-health guideline corpus---a chunk-level relevance benchmark at the
  granularity a RAG system actually retrieves, built around a graded
  clinical-relevance rubric grounded in IR relevance theory, with its judge
  checked against a frontier reference model.
\end{itemize}

Three decisions shape both benchmarks, and they are the through-line of this
paper:
\begin{enumerate}
  \item \textbf{Assemble and filter from expert-authored sources rather than
  author new questions.} The clinical curation already exists in licensing
  exams, physician panels, and published guidelines, so the benchmark's job is
  reliable scoping and uniform packaging, not de-novo clinical authoring.
  \item \textbf{Decompose clinical relevance into a graded, multi-dimensional
  rubric rather than a binary relevant/not-relevant label.} To a midwife, a
  chunk that names a drug and a chunk that gives its dose, route, and threshold
  are not equally useful, and a benchmark that cannot tell them apart cannot
  separate strong retrieval from adequate retrieval.
  \item \textbf{Measure and disclose the limits of the labels}---scope-classifier
  agreement, a frontier-judge check, and a pooling-completeness audit---rather
  than present the benchmark as an oracle.
\end{enumerate}

The companion system paper uses these benchmarks to evaluate the deployed system
and reports those results; this paper documents how the benchmarks were
\emph{built and validated}, and is what the system paper cites for their
provenance. We do not reproduce the deployed-system evaluation here. mamabench
is assembled from external QA sources and is corpus-independent, whereas
mamaretrieval is coupled to a specific corpus chunk set (\S\ref{sec:corpus}).
Section~\ref{sec:limitations} is candid about what each benchmark does and does
not establish; the datasets and construction code are linked in the title
footnote.
