\section{mamaretrieval: A Chunk-Level Relevance Benchmark}
\label{sec:mamaretrieval}

mamaretrieval scores how well a retriever finds the guideline passages that
answer a clinical query, at the chunk granularity a RAG system actually
retrieves. Building it means choosing what to retrieve over (the corpus), what
queries to ask, how to label a (query, chunk) pair's relevance, and how to keep
those labels trustworthy at scale. We start with the corpus, then take query
construction, the relevance rubric, pooling, and judge reliability in turn.

\subsection{The guideline corpus}
\label{sec:corpus}

mamaretrieval is built over the same maternal-health guideline corpus the
deployed system retrieves from, which the companion system
paper~\cite{ren2026mamai} describes in full---its sources, construction, and
packaging. We give here only what the benchmark depends on: the corpus's scale,
its source-quality tiers, and the version coupling that makes the benchmark and
its corpus one artifact.

\paragraph{Scale and sources.}
The corpus bundle (v0.2.0) holds 63{,}650 section-aware chunks drawn from 87
documents spanning international clinical guidance (WHO, NICE, MSF, the
Hesperian midwifery references, and FIGO/EmONC materials), national and local
protocols (the Tanzania and Zanzibar Ministry of Health guidelines), and
standard obstetric and midwifery reference texts. Each chunk carries a canonical
header recording its source, page, and a stable chunk identifier.

\paragraph{Source-quality tiers.}
Sources are rated into quality tiers (from \emph{very high} down to
\emph{low--moderate}) by their clinical depth and relevance to the
Zanzibar OBGYN and midwifery scope. Query construction (below) uses these tiers
as sampling weights, drawing proportionally more from the clinically richest
sources.

\paragraph{Version coupling.}
Chunk identifiers are content-derived---a hash of the chunk text---so they
survive source renames and page drift but change whenever a chunk's text
changes. mamaretrieval references chunks by these identifiers, which couples it
to a specific \emph{chunk set} rather than to a specific embedding. The
distinction is not hypothetical: the deployed system moved from corpus bundle
v0.2.0 to v0.3.0, but that update only re-embedded the corpus---swapping the
on-device embedder---and left all 63{,}650 chunks and their identifiers
untouched, so every mamaretrieval label applies to the deployed v0.3.0 corpus
unchanged. A genuine re-chunking would change the identifiers and require
re-running the labelling pipeline; we therefore treat the
(chunk set, queries, labels) triple as a single versioned artifact rather than
versioning queries and labels on their own.

\subsection{Query construction}
\label{sec:mr-queries}

A retrieval benchmark needs queries, and for this corpus there are none to
mine---no search logs, no question bank tied to these guidelines. We generate
them. Each query in mamaretrieval is written by an LLM from a single corpus
chunk, asking a clinical question that the chunk answers; the chunk that seeded a
query is recorded as a known-relevant anchor for it.

The queries come out of a three-step funnel over the corpus:
\begin{enumerate}
  \item \textbf{Tier-weighted sampling} draws 4{,}540 chunks, with each source's
  quota set by its quality tier (\S\ref{sec:corpus})---the clinically richest
  sources contribute proportionally more, so the query set inherits the corpus's
  quality signal rather than its raw size distribution.
  \item An \textbf{answerability gate} passes each sampled chunk to
  \texttt{Qwen3.6-27B-FP8} (reasoning on, temperature 0), prompted to keep the
  chunk only if a clinician could answer a specific question from it alone and,
  when so, to write exactly that question.\footnote{The full filter-and-generate
  prompt ships with the mamaretrieval construction code (title footnote).}
  \item The kept chunks become the benchmark's \textbf{queries}: each generated
  question gets a stable identifier (\texttt{q\_00001}, \dots), a pointer back to
  its seed chunk, and the source and tier metadata that chunk carried.
\end{enumerate}
The gate keeps 3{,}185 of the 4{,}540 chunks (about 70\%), and the 30\% it
discards are themselves informative: assessment rubrics, learning objectives, and
table-of-contents fragments are rejected because no answerable clinical question
follows from them.

One limitation of this construction is worth flagging up front, because it
shapes which metrics we trust: a query written from a chunk tends to be phrased
like that chunk, which gives dense embedding retrievers an artificial edge over
lexical ones on the very chunk that seeded the query. We do not correct for this
anchor-text effect; we disclose it (\S\ref{sec:mr-judge}).
Section~\ref{sec:limitations} returns to the design's other limits, including
that the queries are clean and each grounded in a single chunk.

\subsection{The graded relevance rubric}
\label{sec:mr-rubric}

With queries and a corpus, the benchmark needs a relevance label for each
(query, chunk) pair: how useful is this chunk for answering this query? Our first
rubric made that judgement binary---relevant or not. It saturated. Take a query
asking for alternative iron and folic-acid dosages (Table~\ref{tab:rubric-example}):
a footnote giving the exact substitute formulation, a dosing section padded with
storage advice, and an adherence-counselling note that never states a dose are,
by a binary test, equally relevant---each is on topic, meaningful, and broadly
actionable. A
benchmark that cannot separate them cannot tell a retriever that surfaces the
dose from one that surfaces the filler.

So we replaced the binary label with a graded one. Each (query, chunk) pair is
scored
\[
  \mathrm{score} = D_1 \times (D_2 + D_3 + D_4) \in \{0,\dots,6\},
\]
one gate and three graded dimensions:
\begin{description}
  \item[$D_1$ --- topic (0/1).] Does the chunk address the query's clinical
  question in the same context---the same timing window (antenatal,
  intrapartum, postpartum)? A zero here forces a score of zero and halts
  scoring, which also keeps the judge from spending tokens on off-topic chunks.
  \item[$D_2$ --- meaningful content (0/1/2).] How much clinical depth the chunk
  carries, from a bare mention to a multi-concept explanation.
  \item[$D_3$ --- actionable guidance (0/1/2).] How specific the guidance is:
  none, a general instruction (``give a uterotonic''), or an exact one
  (``10 IU oxytocin IM'').
  \item[$D_4$ --- density (0/1/2).] What fraction of the chunk is useful for
  \emph{this} query, penalising a relevant passage buried in unrelated text.
\end{description}
A structural rule ties them together: actionable guidance presupposes some
meaningful content ($D_3 > 0 \Rightarrow D_2 > 0$).

This is not an ad-hoc scheme. A topical gate followed by complementary dimensions
that a fixed rule aggregates is the design of the TREC Precision Medicine
track~\cite{roberts2017precision}, and the topical/cognitive/situational split it
rests on is Saracevic's~\cite{saracevic2007relevance}---here $D_1$ is topical and
$D_3$, $D_4$ situational; criteria-based decomposition is also the direction of
recent LLM relevance-judgment work~\cite{farzi2025criteria}. Two further choices
are specific to using an LLM as the judge: an explicit topical gate, and
reasoning-before-verdict---the structured-output schema makes the judge emit its
reasoning before the four scores, as in G-Eval~\cite{liu2023geval}---both
following the now-standard use of a strong LLM as the relevance
assessor~\cite{upadhyay2024umbrela}. The judge is \texttt{Qwen3.5-397B-A17B-FP8}
at temperature 0; \S\ref{sec:mr-judge} reports how well its labels hold up.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.2}
\caption{The graded rubric on one query (\texttt{q\_00041} in release
v0.2.0)---``What are the alternative dosages for iron and folic acid
supplementation?''---as scored by the production judge over four \emph{on-topic}
chunks ($D_1{=}1$ throughout). Text is quoted from the released labels with
markdown list formatting flattened to prose and elisions marked ``\dots'';
sources in brackets (MSF, the MSF obstetric manual; WHO~PCPNC, the WHO
pregnancy/childbirth guide; WHO~ANC, the 2016 WHO antenatal-care guideline). A
binary label cannot separate the first three; the graded score ranks them 6, 4,
and 2. The four chunks (top to bottom) have identifiers \texttt{318a18d9},
\texttt{ad7678f8}, \texttt{466a314b}, and \texttt{27b615ab}.}
\label{tab:rubric-example}
\begin{tabular}{@{}p{0.49\columnwidth}@{\hspace{5pt}}ccccc@{}}
\toprule
Retrieved chunk (verbatim) & $D_1$ & $D_2$ & $D_3$ & $D_4$ & Score \\
\midrule
\emph{[MSF]}~``200 mg ferrous sulfate (65 mg elemental iron) + 400 micrograms
folic acid tablets may be replaced by 185 mg ferrous fumarate (60 mg elemental
iron) + 400 micrograms folic acid tablets.'' & 1 & 2 & 2 & 2 & \textbf{6} \\
\addlinespace
\emph{[WHO PCPNC]}~``Give iron and folic acid. To all pregnant, postpartum and
post-abortion women: routinely once daily in pregnancy and until 3 months after
delivery or abortion; twice daily as treatment for anaemia (double dose). Check
woman's supply \dots\ at each visit and dispense 3 months supply. Advise to store
iron safely \dots'' & 1 & 2 & 1 & 1 & \textbf{4} \\
\addlinespace
\emph{[WHO PCPNC]}~``Motivate on adherence with treatments. Explore local
perceptions about iron treatment (examples of incorrect perceptions: making more
blood will make bleeding worse, iron will cause too large a baby).'' & 1 & 1 & 1
& 0 & \textbf{2} \\
\addlinespace
\emph{[WHO ANC]}~``Resources. Intermittent iron and folic acid supplementation
might cost a little less than daily iron and folic acid supplementation due to
the lower total weekly dose of iron.'' & 1 & 0 & 0 & 0 & \textbf{0} \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:rubric-example} shows the graded rubric on that query. All four
chunks pass the topic gate; the first three also clear a binary rubric's bar
identically---each is meaningful and actionable---yet the graded rubric ranks
them apart. The footnote giving the exact alternative formulation scores 6,
specific and entirely on-query; a dosing section half-taken up by storage and
supply logistics scores 4 ($D_4=1$); and an adherence note that never states a
dose scores 2. The fourth chunk is on topic but pure cost commentary, and drops
to 0---the rubric counts resource text as no meaningful clinical content
($D_2=0$). The judge's reasoning ships with every label and makes such calls
explicit: of that last chunk, it notes the passage sits ``literally under a
\dots\ Resources heading and discusses cost,'' and scores $D_2=0$.

\subsection{Pooling and label completeness}
\label{sec:mr-pooling}

Scoring every chunk against every query is infeasible---63{,}650 chunks times
3{,}185 queries---so mamaretrieval labels a \emph{pool}: for each query, the
union of the top-ranked chunks returned by several retrievers, with every pooled
pair judged and every unpooled chunk treated as not-relevant. This is the
standard information-retrieval pooling protocol, and its one risk is well known:
if the pool misses relevant chunks, the labels under-count them. We measured that
risk rather than assume it away.

\begin{table}[t]
\centering
\small
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.15}
\caption{The six retrievers whose top-$k$ unions form mamaretrieval's label pool,
spanning retrieval paradigms. They \emph{construct} the labels; the
retriever scoreboard \emph{over} them---which one best serves the deployed
system---is reported in the companion paper~\cite{ren2026mamai}, not here. ModernColBERT is the HuggingFace model
\texttt{lightonai/GTE-ModernColBERT-v1}.}
\label{tab:retrievers}
\begin{tabular}{@{}lll@{}}
\toprule
Retriever & Paradigm & Reference \\
\midrule
BM25               & lexical (sparse)      & \cite{robertson2009bm25} \\
MedCPT             & biomedical bi-encoder & \cite{jin2023medcpt} \\
Octen-Embedding-8B & general-purpose dense & model card\tablefootnote{\url{https://huggingface.co/Octen/Octen-Embedding-8B}} \\
voyage-4-large     & frontier API dense    & Voyage~AI\tablefootnote{\url{https://www.voyageai.com}} \\
ModernColBERT      & late-interaction      & \cite{khattab2020colbert} \\
Gecko              & on-device dense       & \cite{lee2024gecko} \\
\bottomrule
\end{tabular}
\end{table}

The first pool took the top-10 chunks from three retrievers (BM25, MedCPT, and
Octen-Em\-bedding-8B): about 24.7 candidates per query and 78{,}571 judged pairs. A
completeness audit on a 100-query sample then measured that pool against a wider
reference (five retrievers at top-20, adding voyage-4-large and ModernColBERT)
and found it had captured only about half of the relevant material (lenient
recall near 0.49): half the relevant chunks were being labelled not-relevant by
omission.

That finding drove the released label set. mamaretrieval v0.2.0 pools the top-20
chunks from six retrievers---the five above plus Gecko---chosen to span retrieval
paradigms so the pool is not biased toward any one
(Table~\ref{tab:retrievers}). Judged across all 3{,}185 queries, this yields
230{,}964 graded labels: the pool the companion system paper draws its label
provenance from. A residual gap remains---chunks that no retriever ranks in its top-20 are still
unjudged---but the same audit bounds it: widening from the initial pool to a
five-retriever top-20 union raised lenient recall from about 0.49 to about 0.99
against its reference, and the released pool adds a sixth retriever, so it is
near-saturated at the top of the ranking. Only absolute coverage-sensitive
numbers need the residual caveat (\S\ref{sec:mr-judge}).

\subsection{Judge reliability and metric guidance}
\label{sec:mr-judge}

A single LLM produces every relevance label, so the benchmark is only as
trustworthy as that judge. We check it against a frontier reference model
(Claude~Opus~4.7) on a 62-pair pilot: the production judge agrees on 59 of 62
pairs at the lenient cut (score $\ge 3$, any useful content) and 53 of 62 at the
strict cut (score $\ge 5$, complete and specific), and a stratified manual
spot-check of its reasoning traces found its calls defensible, including on hard
cases such as drug-name collisions across clinical contexts. Those two
cuts---lenient $\ge 3$ and strict $\ge 5$---are the relevance thresholds the
benchmark reports against.

We are deliberate about what this establishes. It is an agreement check against
another model, not against clinicians: it shows the judge is a reasonable
relevance classifier and that retriever \emph{rankings} are stable under it, but
it does not make any single label ground truth. Each label is one judge's
opinion, and absolute, per-item numbers carry more uncertainty than relative
comparisons.

That uncertainty, together with the pool incompleteness of
\S\ref{sec:mr-pooling}, is why we guide consumers on metric choice. Metrics that
turn on the presence or rank of a relevant chunk---Hit~Rate@$k$ (is any of the
top $k$ relevant?) and MRR (the mean over queries of $1/r$, where $r$ is the rank
of the \emph{first} relevant chunk)---are robust to a pool that misses some
relevant chunks: one hit suffices, and a label missed elsewhere cannot change
that a retrieved relevant chunk is relevant. nDCG@$k$ is
moderately sensitive: it credits every relevant chunk by grade and rank and
normalises against an ideal ranking built from the judged-relevant set, so
missing labels distort both. Precision@$k$ understates any retriever that
surfaces a relevant-but-unjudged chunk (counted as a miss), and Recall@$k$ is the
most exposed, since its denominator---the total count of relevant chunks---is
exactly what pooling under-counts. We recommend leading with the robust metrics
and reading the rest as relative comparisons with the completeness caveat
attached---the order the companion system paper follows when it evaluates
retrievers on mamaretrieval.
