\section{Limitations}
\label{sec:limitations}

We have flagged limits where they arise; here we collect what a user should weigh
before relying on either benchmark.

\paragraph{No human gold standard.}
The scope of mamabench and every relevance label in mamaretrieval are decided by
LLMs, not clinicians. We anchor them externally---cross-model agreement and prior
labels for the scope classifier (\S\ref{sec:mamabench-scope}), a frontier
reference model for the relevance judge (\S\ref{sec:mr-judge})---and we ship
HealthBench's physician labels so a consumer can vet a rubric judge
(\S\ref{sec:mamabench-calibration}). But no midwife reviewed either benchmark at
scale, and those physician labels come from HealthBench's annotator pool, not
midwifery specialists. Each label is best read as one calibrated opinion: model
and retriever \emph{rankings} are robust, absolute per-item numbers less so.

\paragraph{mamaretrieval queries are clean and single-chunk.}
Every query is LLM-written from one chunk and answerable from it
(\S\ref{sec:mr-queries}), with two consequences. First, the benchmark exercises
single-passage retrieval: it does not test questions that can only be answered by
combining several chunks, nor robustness to the noisy, code-switched, or mistyped
queries a hurried nurse might enter. Second, because a query reads like its
source chunk, dense retrievers gain an anchor-text edge that inflates their
absolute scores---another reason to lead with the pooling-robust metrics
(\S\ref{sec:mr-judge}).

\paragraph{The relevance labels are pooled.}
They cover the six-retriever top-20 union; a chunk no retriever surfaced is
unjudged and counted not-relevant (\S\ref{sec:mr-pooling}). Coverage-sensitive
metrics---Recall and Precision---carry that bias, which is why we report
Hit~Rate and MRR first.

\paragraph{Scope, language, and transfer.}
Both benchmarks are English-only and scoped to maternal, neonatal, child, and
reproductive health for a Zanzibar nurse-midwife setting; the numbers need not
transfer to other languages, specialties, or care settings. A few mamabench items
the scope classifier could not converge on are disclosed and have zero net effect
on in-scope counts (\S\ref{sec:mamabench-scope}).
