\section{Introduction}
\label{sec:intro}

Maternal and newborn mortality remain among the highest in sub-Saharan
Africa, and Tanzania---which includes the semi-autonomous archipelago of
Zanzibar---is among ten countries that together account for more than 60\% of
maternal and neonatal deaths worldwide~\cite{bohle2025mamai}. In Zanzibar most
women deliver in government facilities that are often understaffed and
under-resourced, where midwifery care is provided by nurses who lack
midwifery training to international standards~\cite{bohle2025mamai}.
Strengthening midwife-led care is high-leverage---universal access to it
could prevent an estimated 67\% of maternal and 64\% of neonatal
deaths~\cite{nove2021midwives}---yet a basic obstacle is that authoritative,
evidence-based guidance is scarce at the point of care~\cite{swisstph2024mamai}:
the guidelines are long, and looking something up online is unreliable where
connectivity is intermittent and mobile data costly.

We present MAM-AI, a medical question-answering assistant that runs entirely
on an Android device. The nurse-midwife's question is embedded and matched
against an on-device store of guideline passages; the most relevant passages
are added to the prompt; and an on-device language model generates an answer
grounded in those passages, with citations to the source
(Figure~\ref{fig:pipeline}; Section~\ref{sec:system-design}). Because every
stage runs locally, the system needs no connectivity after a one-time setup,
and no patient-related query leaves the device. The generator and embedder
are off-the-shelf open models; our contribution is the system around them,
the curated knowledge base, and the evaluation that determines how it should
be configured.

\begin{figure}[b!]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[
  font=\footnotesize,
  every node/.style={align=center},
  stage/.style={draw=mamteal!60, rounded corners, minimum height=1.05cm,
    minimum width=1.7cm, fill=mamteal!8, inner sep=2pt},
  endpoint/.style={draw=mamamber!75, rounded corners, minimum height=1.05cm,
    minimum width=1.4cm, fill=mamamber!13, inner sep=2pt},
  flow/.style={-{Latex[length=2mm]}, thick, mamteal!70},
]
  \node[endpoint] (q) {\textbf{Query}\\[1pt]\scriptsize clinical\\\scriptsize question};
  \node[stage, right=3.5mm of q] (e) {\textbf{1. Embed}\\[1pt]\scriptsize EmbeddingGemma\\\scriptsize 300M};
  \node[stage, right=3.5mm of e] (r) {\textbf{2. Retrieve}\\[1pt]\scriptsize top-3 chunks};
  \node[stage, right=3.5mm of r] (a) {\textbf{3. Augment}\\[1pt]\scriptsize prompt + context};
  \node[stage, right=3.5mm of a] (g) {\textbf{4. Generate}\\[1pt]\scriptsize Gemma 4 E4B\\\scriptsize int4};
  \node[endpoint, right=3.5mm of g] (ans) {\textbf{Answer}\\[1pt]\scriptsize with citations\\\scriptsize [1] [2] [3]};
  \draw[flow] (q)--(e); \draw[flow] (e)--(r); \draw[flow] (r)--(a);
  \draw[flow] (a)--(g); \draw[flow] (g)--(ans);
  \node[draw=mamteal!55, dashed, rounded corners, fill=mamteal!6,
    below=9mm of r, minimum width=6cm, inner sep=4pt] (kb)
    {\textbf{On-device knowledge base}\\[1pt]\scriptsize
     87 guideline documents \,\textbullet\, 63{,}650 chunks
     \,\textbullet\, SQLite vector store};
  \draw[flow, dashed] (kb)--(r);
  \begin{scope}[on background layer]
    \node[draw=mamamber!60, dotted, thick, rounded corners, inner sep=5mm,
      fit=(q)(ans)(kb),
      label={[font=\scriptsize\itshape, text=mamamber]above:Runs 100\% on-device --- fully offline}] {};
  \end{scope}
\end{tikzpicture}%
}
\caption{The on-device RAG pipeline. The query is embedded with
EmbeddingGemma and matched against an on-device guideline corpus; the top-3
passages are added to the prompt before Gemma~4~E4B generates a cited
answer. Every stage runs on the device.}
\label{fig:pipeline}
\end{figure}

Safety is the governing constraint throughout---wrong advice can cause direct
harm---and we are explicit that MAM-AI is a thoroughly evaluated research
prototype, not a fielded product: a planned field test could not proceed this
cycle, so we have no data on how the system performs in real clinical use;
all our evidence comes from benchmarks and LLM-judge scoring.

The evaluation is layered---end-to-end, then retriever and generator in
isolation, then latency---and yields three findings that structure the paper:

\begin{enumerate}
  \item \textbf{A redesigned system prompt is a major lever on answer
  quality.} The base model deflected (``see a doctor'') on about a third of
  questions; a redesigned prompt cut that to roughly 3\% and about doubled the
  correct guidance that answers carry, with almost no increase in unsafe
  answers and no genuine dangerous ones.

  \item \textbf{On-device retrieval is strong, but \emph{ranking} is not the
  lever.} A 300M on-device embedder retrieves nearly as well as cloud
  systems---third of seven retrievers, single-digit points behind the best---yet
  adding its context to the small generator is net-neutral to slightly
  negative on average, and neither a better embedder nor a fine-tuned reranker
  changes that. Two things do govern answer quality: how well the small
  generator uses the context it is handed, and---decisively, per query---the
  coverage, specificity, and curation of the corpus, since a small faithful
  model is only as good as the chunk it receives. The remaining headroom is in
  the generator and the corpus, not in re-ranking.

  \item \textbf{The generator matters most, and faithfulness decides which
  model we deploy.} Model capability dominates the results: a size ladder shows
  open-ended quality rising steeply with generator size---roughly quintupling
  on HealthBench from the deployed 4B to a 30B server model, then plateauing
  toward the frontier---while quantization precision and retrieval barely move
  it. Between the two same-size on-device candidates we then choose on
  faithfulness: we deploy the one that is as faithful to its sources as a
  frontier model and about twice as faithful as the other, more-helpful but
  less safe, candidate. Deployment is thus a safety-versus-usefulness choice.
\end{enumerate}

\paragraph{Contributions.}
\begin{itemize}
  \item A deployable on-device medical RAG system for nurse-midwives in a
  low-connectivity setting, running fully offline on commodity Android
  hardware (Section~\ref{sec:system-design}).
  \item A layered evaluation---end-to-end, retriever, generator, and
  latency---on two purpose-made benchmarks and a validated LLM judge, applied
  to the exact deployed configuration, which yields the three findings above
  (Sections~\ref{sec:methodology}--\ref{sec:latency}).
\end{itemize}

\paragraph{Availability.}
All components are open source and the deployed configuration has a public
browser demo: Figure~\ref{fig:artifacts} maps how they fit together and
Table~\ref{tab:artifacts} links each one. The two benchmarks are described in
detail in a companion paper~\cite{yi2026mamabench}.

\begin{table}[t]
\centering
\footnotesize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.18}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Component} & \textbf{Repository / resource} \\
\midrule
App \& on-device RAG pipeline & \url{https://github.com/nmrenyi/mamai} \\
Knowledge base \& guideline bundle & \url{https://github.com/nmrenyi/mamai-medical-guidelines} \\
\texttt{mamabench} (QA benchmark) & \url{https://huggingface.co/datasets/nmrenyi/mamabench} \\
\texttt{mamabench} (construction code) & \url{https://github.com/nmrenyi/mamabench} \\
\texttt{mamaretrieval} (retrieval benchmark) & \url{https://huggingface.co/datasets/nmrenyi/mamaretrieval} \\
\texttt{mamaretrieval} (construction code) & \url{https://github.com/nmrenyi/mamaretrieval} \\
Evaluation harness & \url{https://github.com/nmrenyi/mamai-eval} \\
Browser demo (deployed config) & \url{https://nmrenyi-mamai-demo.hf.space} \\
\bottomrule
\end{tabular}
\caption{The released MAM-AI components and where to find them; the two
benchmarks are published both as Hugging Face datasets and as GitHub
repositories. Figure~\ref{fig:artifacts} shows how the pieces relate.}
\label{tab:artifacts}
\end{table}

\begin{figure}[t]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[
  font=\footnotesize,
  every node/.style={align=center},
  repo/.style={draw=mamteal!70, rounded corners, fill=mamteal!8,
    inner sep=3pt, text width=2.9cm, minimum height=0.95cm},
  flow/.style={-{Latex[length=2mm]}, thick, mamteal!75},
  lbl/.style={font=\scriptsize\itshape, text=black!65, inner sep=1.5pt},
]
  % --- LEFT column: the product pipeline, stacked top-to-bottom ---
  \node[repo] (kb) at (0,0) {\texttt{\textbf{mamai-medical-}}\\\texttt{\textbf{guidelines}}\\[1pt]
    \scriptsize knowledge-base pipeline};
  \node[repo] (app) at (0,-2.5) {\texttt{\textbf{mamai}}\\[1pt]
    \scriptsize on-device app \& RAG};
  \node[repo] (demo) at (0,-5.0) {\texttt{\textbf{mamai-demo}}\\[1pt]
    \scriptsize browser demo \emph{(HF\,Space)}};
  \draw[flow] (kb) -- node[lbl,right]{corpus bundle\\(v0.3.0)} (app);
  \draw[flow] (app) -- node[lbl,right]{deployed\\config} (demo);
  % --- RIGHT: evaluation harness on top; the two benchmarks stacked compactly
  %     below it, grouped as one block that feeds the harness ---
  \node[repo] (eval) at (5.6,-2.5) {\texttt{\textbf{mamai-eval}}\\[1pt]
    \scriptsize evaluation harness};
  \node[repo] (b1) at (5.6,-4.55) {\texttt{\textbf{mamabench}}\\[1pt]
    \scriptsize QA benchmark};
  \node[repo] (b2) at (5.6,-5.95) {\texttt{\textbf{mamaretrieval}}\\[1pt]
    \scriptsize retrieval benchmark};
  % group the two benchmarks; a single arrow from the block feeds the harness
  \begin{scope}[on background layer]
    \node[draw=mamamber!60, dashed, rounded corners, fill=mamamber!7,
      inner sep=2mm, fit=(b1)(b2)] (benchbox) {};
  \end{scope}
  \draw[flow] (benchbox.north) -- node[lbl,right]{test data} (eval.south);
  \draw[flow] (eval) -- node[lbl,above]{evaluates} (app);
  \node[font=\scriptsize\itshape, text=mamamber, align=center,
    below=0.5mm of benchbox] {documented in the\\companion paper~\cite{yi2026mamabench}};
\end{tikzpicture}%
}
\caption{The MAM-AI open-source ecosystem. Boxes are public GitHub repositories
under the \texttt{nmrenyi} organisation. \texttt{mamai-medical-guidelines}
builds the versioned guideline bundle that the \texttt{mamai} app retrieves over
on-device; \texttt{mamai-demo} runs the same deployed configuration in the
browser. The \texttt{mamabench} and \texttt{mamaretrieval} benchmarks (also
released as Hugging Face datasets) drive \texttt{mamai-eval}, which scores the
deployed system---the evaluation this paper reports. The benchmarks themselves
are documented in the companion paper.}
\label{fig:artifacts}
\end{figure}
