\section{System Design}
\label{sec:system-design}

% Deliberately high-level. Low-level app mechanics (Flutter UI, platform
% channels, lazy init, stream/job lifecycle) are omitted -- they are
% readable from the open-source repo and not needed to interpret results.
% Keep only what is load-bearing for the evaluation: the RAG flow, the two
% deployed models, the system prompt (a designed artifact whose effect is a
% headline result), and the deployment choices that have measurable
% consequences (notably the GPU FP16 default and its long-context failure).
% Verified 2026-06-25 against mamai/config/* and CLAUDE.md.

MAM-AI is an Android application that answers a nurse-midwife's clinical
question entirely on-device: embedding, retrieval, and generation all run on
the device, and no query ever leaves it. This is forced by the deployment
context---health facilities in Zanzibar cannot assume connectivity, and data
costs make online tools unreliable---and it has the side benefit that
patient-related queries are never transmitted off the device.

\subsection{On-Device RAG Pipeline}

The system is a retrieval-augmented generation (RAG) pipeline
(Figure~\ref{fig:pipeline}). The user's question is embedded and matched
against an on-device vector store of guideline passages; the top-3 passages
are injected into the prompt alongside the question; and the generator
produces a streamed answer with inline citations back to the source
passages. Every stage runs locally against a pre-computed, versioned
knowledge base (Section~\ref{sec:corpus}); the application is fully offline
after a one-time setup (models and corpus are downloaded or sideloaded onto
the device).

\subsection{Models}

Both model components are off-the-shelf open models from Google, run through
the LiteRT-LM on-device runtime; what is contributed here is the system, the
knowledge base, and the evaluation built around them.

\begin{itemize}
  \item \textbf{Generator: Gemma~4~E4B} (int4-quantized, 3.66\,GB). Decodes
  the final answer from the question and retrieved passages.
  \item \textbf{Retriever: EmbeddingGemma-300M} (768-dimensional, quantized).
  Embeds the query for vector search. It replaced an earlier on-device
  embedder (Gecko); the retrieval gain from that swap, and why it does not
  translate into better answers, are quantified in
  Section~\ref{sec:retrieval-results}.
\end{itemize}

\subsection{The System Prompt}
\label{sec:system-prompt}

The deployed system prompt (``G1'') is a designed artifact, not boilerplate,
and its effect is one of the paper's headline results
(Section~\ref{sec:endtoend}). The base prompt left the model evasive---on
roughly a third of questions it deflected to ``see a doctor'' rather than
giving first-line guidance. G1 targets that failure with a set of explicit
levers: it addresses the user as a clinician rather than the patient, treats
any clinical question (including emergencies and ones naming a location) as
in scope, requires concrete first-line management \emph{before} advising
referral, permits naming standard first-line drugs and doses with a
formulary caveat, and prefers locally available options. The prompt is
English-only; an earlier bilingual (English/Swahili) configuration was
withdrawn because a safety-critical prompt cannot ship in a language it was
not evaluated in (Section~\ref{sec:limitations}).

\subsection{Deployment Configuration}

Retrieval depth is fixed at top-3 and the context window at 4{,}096 tokens.
The generator defaults to CPU---the broadly compatible, safer backend---with
GPU execution as an opt-in: it is faster where a compatible GPU is present
(\S\ref{sec:latency}) but has a higher hardware bar, so it is not the shipped
default and falls back to CPU on failure. The GPU path matters for one non-obvious reason: it
runs FP16 attention by default, which decodes correctly at the deployed
depth but \emph{degrades deterministically past roughly 5{,}000 tokens of
context}---the 4{,}096-token ceiling is a safety margin below that cliff
(characterized in Section~\ref{sec:latency}). The knowledge base ships as a
checksum-pinned bundle (v0.3.0), so a given build retrieves from a bit-exact,
auditable corpus.
