\section{Knowledge Base}
\label{sec:corpus}

% Placed right after System Design: the full system, data included, is
% described before it is evaluated. (Deviates from the webpage, which puts
% the knowledge base after the evaluation -- burying it there undersold it.)
% Copyright posture (per project decision): the surface here is neutral ---
% name the openly-available authoritative sources, group reference texts
% generically, no permission/licensing admission that would spotlight the
% issue. Full per-document provenance lives in the open bundle manifest for
% researchers (open science). Do NOT print verbatim excerpts from commercial
% texts; the sample chunk is from a freely-available WHO guideline.

The system's answers are only as good as what it can retrieve, so the
knowledge base is its foundation. It is a curated corpus of 87 authoritative
clinical guideline documents---spanning maternal, newborn, OBGYN, and
reproductive health---processed into 63{,}650 passages and embedded for
fully offline on-device search (Figure~\ref{fig:pipeline}). The resulting
on-device store is about 275\,MB.

\subsection{Sources}

The corpus draws on authoritative international guidelines, field references
written for low-resource settings, and Zanzibar and Tanzania national
materials (Table~\ref{tab:sources}).

\begin{table}[t]
\caption{Source families in the knowledge base. Per-document provenance and a
SHA-256 checksum for every source are recorded in the bundle manifest.}
\label{tab:sources}
\small
\begin{tabular}{@{}p{0.26\columnwidth}p{0.66\columnwidth}@{}}
\toprule
\textbf{Source family} & \textbf{Coverage} \\
\midrule
WHO & Antenatal, intrapartum, postnatal, and newborn care; postpartum
haemorrhage, eclampsia, obstructed-labour, and sepsis modules; abortion,
contraception, STI, and gender-based-violence care \\[2pt]
NICE (UK) & Antenatal, intrapartum, and postnatal care; perinatal mental
health; contraception \\[2pt]
ICM, RCM, RCOG, ACOG & International and royal-college midwifery competencies
and clinical guidance \\[2pt]
Hesperian, MSF & Field references for low-resource settings: community
midwifery, essential obstetric and newborn care \\[2pt]
Zanzibar / Tanzania national & Scope of practice, Ministry-of-Health
competency standards, midwifery training curricula, and service standards \\[2pt]
Assessment and reference & UK NMC competence-test materials,
newborn-resuscitation protocols, and standard midwifery reference texts \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Construction and Packaging}

PDFs are converted with an ML layout model (marker-pdf) that recovers tables
and headings, then chunked \emph{structure-first}: heading boundaries, not
page breaks, define passage boundaries, and each chunk carries a
parent-section breadcrumb so it stays self-contained when retrieved. Every
chunk receives a content-hash identifier (the SHA-256 of its text,
truncated), which is stable across edits and doubles as a citation key. The
chunks are embedded with EmbeddingGemma and stored in a SQLite vector store,
shipped as a single versioned bundle (v0.3.0) with a SHA-256 checksum for
every document; each retrieved passage is therefore traceable to an exact
source file. The full construction pipeline is open
source.\footnote{\url{https://github.com/nmrenyi/mamai-medical-guidelines}}

Each stored passage carries a \texttt{[SOURCE | PAGE | CID]} header, which is
what lets the system cite an answer back to a specific guideline page. For
example:

\begin{lstlisting}[style=prompt]
[SOURCE: WHO_Complications_2017 | PAGE: 204 | CID: c5c7acd564ff3c8b]
> Hypertensive disorders of pregnancy > Magnesium sulfate maintenance dose

Withhold or delay the drug if:
- respiratory rate falls below 16 breaths per minute;
- patellar reflexes are absent;
- urinary output falls below 30 mL per hour over the preceding four hours.
\end{lstlisting}

The corpus is treated as a fixed input in this report; it was re-embedded
(not changed) when the retriever was upgraded, and it was not expanded this
cycle, so its coverage is a stated limitation (Section~\ref{sec:limitations})
rather than a result.
