\section{Evaluation Methodology}
\label{sec:methodology}

We evaluate the \emph{exact deployed configuration}---Gemma~4~E4B with the G1
prompt, EmbeddingGemma top-3 retrieval, and corpus bundle v0.3.0---rather
than a research stand-in. This section fixes the framework, the benchmarks,
the judge, and the metrics that the results sections
(\S\ref{sec:endtoend}--\S\ref{sec:latency}) report against.

\subsection{A Layered Evaluation}

We evaluate in layers: the end-to-end system first (\S\ref{sec:endtoend}),
then the retriever (\S\ref{sec:retrieval-results}) and the generator
(\S\ref{sec:generator}) in isolation, then on-device latency
(\S\ref{sec:latency}). The components are built bottom-up---a weak retriever
would make generator evaluation meaningless, and an end-to-end score alone
confounds retrieval quality with generation faithfulness---but presented
top-down: we lead with the deployed system's behaviour, then decompose it to
locate the cause. Except for the latency measurements, which require the
phone, evaluation runs on a cluster; \S\ref{sec:calibration} shows this is a
faithful proxy for on-device behaviour.

\subsection{Benchmarks}

Two purpose-built benchmarks supply the test data; both are documented in a
companion paper, so we describe them only as far as the results require.
\textbf{mamabench}\footnote{mamabench --- dataset:
\url{https://huggingface.co/datasets/nmrenyi/mamabench}; construction code:
\url{https://github.com/nmrenyi/mamabench}.} is a maternal/neonatal/OBGYN
question-answering benchmark (25{,}949 items across multiple-choice,
open-ended, and rubric-graded tracks). The end-to-end evaluation is anchored on its two open-ended tracks:
the \emph{Kenya Clinical Vignettes} (312 nurse-written primary-care cases,
close to the deployment setting) and \emph{HealthBench-oss} (1{,}209 items
with physician-written rubrics); multiple-choice serves only as a control.
\textbf{mamaretrieval}\footnote{mamaretrieval --- dataset:
\url{https://huggingface.co/datasets/nmrenyi/mamaretrieval}; construction
code: \url{https://github.com/nmrenyi/mamaretrieval}.} pairs 3{,}185 clinical
queries with graded ($0$--$6$) relevance labels over the guideline corpus,
and is used to score the retriever (\S\ref{sec:retrieval-results}).

\subsection{Metrics}

Three metrics recur across the open-ended results and are defined once here;
retrieval and faithfulness metrics are defined where they are used
(\S\ref{sec:retrieval-results}, \S\ref{sec:generator}).
\begin{itemize}
  \item \textbf{Key-fact recall} --- the fraction of an expert reference
  answer's key facts the response conveys, counting partial mentions at half
  weight. It scores the \emph{short-answer-question} (SAQ) tracks---free-text
  questions with a gold answer, as distinct from the rubric-graded HealthBench
  track. (Each SAQ item ships with a physician-written list of key facts, and
  recall is the share of that list the answer states.)
  \item \textbf{Safety} --- a four-level judgement by \emph{harm pathway},
  not tone: \emph{safe}, \emph{minor concern}, \emph{potentially harmful}
  (a small, often catchable harm path), and \emph{dangerous} (would harm if
  followed at face value).
  \item \textbf{\texttt{weighted\_met}} --- the HealthBench rubric score:
  credit for satisfied criteria, net of penalties for violated ``should-not''
  criteria, normalised by the achievable positive credit. It can go negative,
  so we report it decomposed into positive credit and penalty incurred.
\end{itemize}

\subsection{Validating the Judge}
\label{sec:judge}

The open-ended tracks are scored by an LLM judge, so we first validate the
judge against physicians. Five candidates run the HealthBench rubric grader
over the same 6{,}853 physician-labelled (prompt, completion, criterion)
triples, and we compare each judge's labels to the physicians'. The headline
calibration axis is the \emph{met-rate}---how often a judge labels a
criterion satisfied---whose target is the physician rate of 77\%. The judges
fall on a clean spectrum around that target (Figure~\ref{fig:judge}):
stronger closed-source models under-rate (gpt-5 by nearly 19 points), while
the largest open models over-rate.

\begin{figure}[t]
\centering
\scalebox{1.2}{%
\begin{tikzpicture}[font=\footnotesize]
  % met-rate 50%..90% mapped at 0.18 cm per point: x = (v-50)*0.18
  \draw[black!55] (0,0) -- (7.2,0);
  \foreach \v/\x in {50/0, 60/1.8, 70/3.6, 80/5.4, 90/7.2}{
    \draw[black!55] (\x,0) -- (\x,-0.08);
    \node[below=1pt, font=\scriptsize, text=black!60] at (\x,-0.08) {\v\%};
  }
  % physician target at 77% -> x=4.86
  \draw[mamamber, dashed, thick] (4.86,-0.12) -- (4.86,2.0);
  \node[font=\scriptsize, text=mamamber, align=center] at (4.86,2.3)
    {physician 77\%\\(target)};
  % other judges: dot + connector + label
  \foreach \x/\nm/\val in {1.49/gpt-5/58.3, 3.6/gpt-5-mini/70.0,
                           5.99/Nemotron/83.3, 6.98/Maverick/88.8}{
    \fill[black!55] (\x,0) circle (2.2pt);
    \draw[black!30] (\x,0.05) -- (\x,0.66);
    \node[font=\scriptsize, align=center] at (\x,0.92) {\nm\\\scriptsize \val\%};
  }
  % deployed judge highlighted: gpt-oss-120b at 73.2% -> x=4.18
  \fill[mamteal] (4.18,0) circle (3pt);
  \draw[mamteal!50] (4.18,0.05) -- (4.18,1.2);
  \node[font=\scriptsize, text=mamteal, align=center] at (4.18,1.46)
    {\textbf{gpt-oss-120b}\\\scriptsize\textbf{73.2\% (deployed)}};
  % under/over annotations
  \node[font=\scriptsize\itshape, text=black!55] at (1.3,-0.62) {$\leftarrow$ under-rates};
  \node[font=\scriptsize\itshape, text=black!55] at (6.2,-0.62) {over-rates $\rightarrow$};
\end{tikzpicture}%
}
\caption{Judge calibration spectrum: each candidate's met-rate against the
physician target (77\%). Stronger closed-source judges under-rate; larger
open models over-rate; the deployed \texttt{gpt-oss-120b} sits closest to
physicians.}
\label{fig:judge}
\end{figure}

Met-rate alone is not enough, because overall agreement is misleading: with
most criteria satisfied, a judge that simply says ``met'' scores well. The
diagnostic is the per-class split (Table~\ref{tab:judge})---agreement on
physician-MET versus physician-NOT-MET rows. Lenient judges score
near-perfectly on MET rows but collapse on NOT-MET rows (the larger open
models, and the trivial always-``met'' baseline, all show this signature);
over-strict judges do the reverse.

We deploy \texttt{gpt-oss-120b} for two reasons:
\begin{itemize}
  \item \textbf{Calibration.} Its met-rate is the closest to physicians
  ($-3.8$ points), and it sits well clear of the rubber-stamping models on
  NOT-MET rows---the best overall balance, since escalating to the stronger
  closed-source judges only worsens calibration (gpt-5 under-rates by nearly
  19 points).
  \item \textbf{Open and reproducible.} It is an open model we host on our
  own GPU, so the full rubric scoring is reproducible and free to re-run;
  the closed-source candidates bill per call and cannot be reproduced
  independently.
\end{itemize}
Its mild conservative bias is a known, disclosed direction on every rubric
score that follows.

\begin{table}[t]
\caption{Judge--physician agreement (\%) on the 6{,}853-triple set, the
rubber-stamp detector. \textbf{Cons.}: agreement with the physician
\emph{consensus} label, on the rows where physicians agreed (overall
accuracy). \textbf{MET}: of the criteria physicians marked \emph{met}, the
fraction the judge also marks met. \textbf{NOT-MET}: of the criteria
physicians marked \emph{not met}, the fraction the judge also marks not
met---the decisive column, since a lenient judge scores high on MET but low
here. The deployed \texttt{gpt-oss-120b} stays balanced across both classes;
the trivial baselines max out one class at the cost of the other.}
\label{tab:judge}
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
Judge & Cons. & MET & NOT-MET \\
\midrule
Always ``met'' (trivial) & 85.2 & 100\phantom{.0} & 0\phantom{.0} \\
Llama-4 Maverick & 88.7 & 96.3 & 44.6 \\
Nemotron-253B & 87.4 & 92.9 & 55.3 \\
\textbf{gpt-oss-120b (deployed)} & 81.6 & 84.0 & 67.9 \\
gpt-5-mini & 79.1 & 80.7 & 69.5 \\
gpt-5 & 71.8 & 70.1 & 81.9 \\
Always ``not met'' (trivial) & 14.8 & 0\phantom{.0} & 100\phantom{.0} \\
\bottomrule
\end{tabular}
\end{table}

\FloatBarrier
\subsection{Device--Cluster Fidelity}
\label{sec:calibration}

The deployed app runs on the phone, but for efficiency the open-ended
evaluation is generated on a GPU cluster---a different runtime on different
hardware (LiteRT-LM at FP16 on the Android GPU vs.\ a 4-bit Q4\_0
quantisation under llama.cpp on an A100). Before trusting cluster scores as a
stand-in for the device, we re-ran the open-ended questions on the real
device and scored both stacks with the same judge (SAQ in full, 369 per arm;
HealthBench on a 150-per-arm stratified sample, as the full set is
$\sim$12\,h per arm on-device). We compare on the no-RAG arm, which feeds
both stacks \emph{identical} inputs and so isolates the runtime gap; the
deployed +RAG arm would additionally reflect the device's own retrieval and
no longer be a clean stack comparison.

\begin{table}[t]
\caption{Open-ended device--cluster fidelity (no-RAG, identical inputs): the
deployed device (LiteRT-LM) vs.\ the cluster (llama.cpp Q4\_0) on the same
questions, same judge. Recall and \texttt{weighted\_met} $\uparrow$ better,
harm $\downarrow$ better. SAQ datasets: Kenya, AfriMedQA-SAQ (AfriMed),
WHB ($n{=}20$); HealthBench subsets (HB): oss\_eval, consensus, hard. The
device matches or beats the cluster on every metric.}
\label{tab:calib}
\centering
\setlength{\tabcolsep}{4pt}
\small
\begin{tabular}{@{}lccccccccc@{}}
\toprule
& \multicolumn{3}{c}{Recall $\uparrow$ (SAQ)}
& \multicolumn{3}{c}{Harm \% $\downarrow$ (SAQ)}
& \multicolumn{3}{c}{\texttt{weighted\_met} $\uparrow$ (HB)} \\
\cmidrule(lr){2-4}\cmidrule(lr){5-7}\cmidrule(lr){8-10}
& Kenya & AfriMed & WHB & Kenya & AfriMed & WHB & oss & cons.\ & hard \\
\midrule
Device  & 0.194 & 0.211 & 0.079 & 19.2 & 8.1 & 5.0 & 0.045 & 0.657 & $-0.137$ \\
Cluster & 0.178 & 0.164 & 0.039 & 20.8 & 16.2 & 20.0 & $-0.036$ & 0.573 & $-0.168$ \\
\bottomrule
\end{tabular}
\end{table}

Across both tracks the device matches or slightly exceeds the cluster on
every metric (Table~\ref{tab:calib}): higher key-fact recall and lower harm
on all three SAQ datasets, and higher \texttt{weighted\_met} on all three
HealthBench subsets. The deployed device is therefore at least as good as the
cluster numbers we report, which stand as a mildly pessimistic lower bound
(the 4-bit cluster quantisation is slightly lossier than the on-device
bundle), so every end-to-end conclusion holds on-device. In one respect the two
differ: on the cluster, the model's internal reasoning occasionally leaked
into the scored answer---an artifact of the cluster's prompt template---
whereas this never occurred across the 669 on-device generations, so it is an
evaluation quirk, not model behaviour. The same judge scores both stacks, so
its bias cancels in the delta; per-subset samples are small (especially WHB),
so single-subset deltas are indicative rather than precise.

A final caveat: this check fixes a single device and backend (the on-device
GPU; \S\ref{sec:latency}). The deployed model uses the same 4-bit weights on
every phone, but the compute around them is not fixed---on-device behaviour
can still shift with the hardware backend (CPU vs.\ GPU) and its numeric
precision (e.g.\ the GPU's FP16 vs.\ FP32 attention) across different
SoCs---so another phone need not reproduce these numbers exactly. The
cluster, a single fixed environment, is by contrast straightforward to
reproduce, a further reason to anchor the evaluation there.
