\section{End-to-End Evaluation}
\label{sec:endtoend}

This section evaluates the whole system as a clinician would meet it. To
place the deployed choice in context we run a $3\times3$ matrix: the two
deployable on-device generators (Gemma~4~E4B and Gemma~3n~E4B) plus an
unconstrained frontier model (Qwen3.5-397B) as a capability ceiling, each
under three prompts---the plain baseline, \mbox{+G1} (the
deflection/scope-fix prompt of \S\ref{sec:system-prompt}), and \mbox{+G1+G2}
(G1 plus a consultation-workflow structure). Every cell uses the deployed
RAG pipeline (EmbeddingGemma top-3, bundle v0.3.0) and is scored on the two
open-ended tracks---Kenya vignettes and HealthBench-oss---by the validated
gpt-oss-120b judge. Table~\ref{tab:matrix} is the result; the same matrix
under oracle context, which isolates faithfulness, is analysed in
\S\ref{sec:generator}.

\begin{table}[t]
\caption{End-to-end matrix (the deployed +RAG configuration): three
generators $\times$ three prompts on Kenya vignettes ($n=312$) and
HealthBench-oss ($n\approx1{,}209$), gpt-oss-120b judge. Recall and
\texttt{w\_met} (HealthBench \texttt{weighted\_met}) higher is better;
deflection, harm (potentially-harmful + dangerous), and dangerous count
(of 312) lower is better. The deployed cell is Gemma~4~+G1 (bold). Oracle
faithfulness for the same matrix is in \S\ref{sec:generator}.}
\label{tab:matrix}
\centering
\small
\begin{tabular}{@{}lrrrrr@{}}
\toprule
 & \multicolumn{4}{c}{Kenya vignettes} & HB \\
\cmidrule(lr){2-5}\cmidrule(lr){6-6}
Generator $\cdot$ prompt & Recall & Defl. & Harm & Dang. & \texttt{w\_met} \\
\midrule
\multicolumn{6}{@{}l}{\textit{Gemma 4 E4B (deployed generator)}} \\
\quad baseline & 0.139 & 32.7\% & 12.8\% & 0 & 0.000 \\
\quad \textbf{+G1 (deployed)} & \textbf{0.279} & \textbf{3.2\%} & \textbf{16.0\%} & \textbf{1} & \textbf{0.038} \\
\quad +G1+G2 & 0.338 & 1.6\% & 17.0\% & 0 & 0.052 \\
\midrule
\multicolumn{6}{@{}l}{\textit{Gemma 3n E4B}} \\
\quad baseline & 0.297 & 1.9\% & 25.6\% & 4 & 0.083 \\
\quad +G1 & 0.351 & 0.6\% & 34.9\% & 11 & 0.110 \\
\quad +G1+G2 & 0.389 & 0.0\% & 33.7\% & 15 & 0.119 \\
\midrule
\multicolumn{6}{@{}l}{\textit{Qwen3.5-397B (frontier ceiling)}} \\
\quad baseline & 0.295 & 0.3\% & 8.6\% & 1 & 0.142 \\
\quad +G1 & 0.420 & 0.0\% & 12.5\% & 0 & 0.161 \\
\quad +G1+G2 & 0.553 & 0.0\% & 4.2\% & 0 & 0.245 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The prompt fixes deflection on the deployed model}

The baseline Gemma~4 is \emph{safe but useless}: it deflects on 32.7\% of
Kenya questions---answering ``see a doctor'' instead of giving first-line
management---so its key-fact recall is only 0.139 and its HealthBench
\texttt{weighted\_met} is essentially zero. It rarely says anything wrong
because it rarely says anything. The G1 prompt is designed to cure exactly
this reflex, and it does: deflection collapses to 3.2\%, recall roughly
doubles to 0.279, and \texttt{weighted\_met} rises off the floor---at almost
no safety cost (dangerous answers $0\to1$, harm $12.8\to16.0\%$). Adding the
G1+G2 consultation structure lifts recall further to 0.338. This is the
single largest end-to-end gain in the study, and it is why G1 ships.

\subsection{Helpfulness versus safety: why Gemma 4, not Gemma 3n}

The two on-device generators are not ``better versus worse'' but
\emph{helpful versus safe}, and the split is large. Gemma~3n is the more
helpful model---higher recall at every prompt (0.30--0.39 vs.\ 0.14--0.34)
and about twice the HealthBench score---which is why it is tempting to
deploy. But every safety and faithfulness signal favours Gemma~4 by roughly
$2\times$: harm rate (12.8--17.0\% vs.\ 25.6--34.9\%) and, most sharply, the
count of \emph{dangerous} answers---ones that would harm a patient if
followed---which is $0$/$1$/$0$ for Gemma~4 across the three prompts but
$4$/$11$/$15$ for Gemma~3n. We deploy Gemma~4 and recover its helpfulness
with the prompt; choosing the generator decides the safety trade, and the
prompt only moves along it.

\subsection{Dangerous-case adjudication}

Because ``dangerous'' is the decisive axis, we hand-read the flagged cases
with a clinician's eye. Gemma~3n's are genuine, serious errors, concentrated
on the prompt lever that authorises naming drugs and doses: oxytocin dosed in
``mg/min'' rather than milli-units (a $\sim$1000$\times$ overdose),
ceftriaxone in a two-week-old neonate (contraindicated; kernicterus risk),
benzathine penicillin at 250{,}000~units for syphilis (a $\sim$10$\times$
underdose), and telling a clinician that an HIV-negative assaulted child is
``not eligible for PEP'' (backwards). On a non-deflecting weak grounder, the
deflection-fix prompts buy recall by spending safety. Gemma~4's single
flagged case (under +G1) is, by contrast, a judge over-flag---standard
choking first-aid scored as dangerous---consistent with its near-zero genuine
rate.

\subsection{HealthBench, decomposed: the gap is completeness, not harm}

HealthBench's \texttt{weighted\_met} nets reward criteria against
penalty (``should-not'') criteria, so we split the two. The positive
score---completeness, how much required guidance the answer contains---rises
monotonically with model capability (Gemma~4 0.18--0.23, Gemma~3n
0.26--0.29, Qwen 0.31--0.41), and both prompts raise it within each tier. The
penalty rate is by contrast strikingly flat ($\sim$0.34--0.39) across all
nine cells: on this rubric's negative axis neither generator nor prompt moves
active harm much. So the \texttt{weighted\_met} ordering is a
\emph{completeness} ranking, not a safety one---which is why the ``3n is less
safe'' signal must be read from the targeted Kenya safety probe (the
dangerous count) and oracle faithfulness (\S\ref{sec:generator}), not from
HealthBench penalties.

\subsection{The capability ceiling: the lever is grounding, not the prompt}

The Qwen-397B row shows what the same prompt does as capability rises. The
identical +G1+G2 prompt \emph{hurts} Gemma~3n (dangerous $4\to15$), is
\emph{neutral} on Gemma~4 ($0\to0$), and is \emph{pure upside} on Qwen
($1\to0$ dangerous while recall climbs to 0.553)---the best cell in the
matrix on every axis at once (recall 0.553, harm 4.2\%, zero dangerous). The
prompts were never the problem; what varies is whether the generator can act
on a request for more clinical substance \emph{without} inventing
unsupported claims. That reframes the on-device goal: the lever is
generator-side RAG-grounding---moving the small models toward the Qwen
corner---not further prompt tuning or retrieval swaps (\S\ref{sec:discussion}).

\subsection{Multiple-choice as a negative control}

On multiple-choice the deployed model shows the opposite of the open-ended
result: across 23{,}241 MCQ items, adding RAG \emph{lowers} accuracy from
53.7\% to 51.9\% ($-1.8$ points). This is not a deployment failure but a
domain mismatch---the maternal/neonatal corpus is off-topic for broad
USMLE-style questions, so retrieved chunks distract more than they help. It
confirms that the deployment-relevant signal is the open-ended tracks above,
where retrieval has on-topic material to contribute, not multiple-choice.

\subsection{Model Size [PENDING -- new experiment]}
% TODO (results announced 2026-06-25, not yet run): a model-size ablation to
% quantify how generator size drives quality/faithfulness, strengthening
% finding 3 ("the generator matters most"). Likely a size sweep (e.g., Gemma
% E2B vs E4B, and/or intermediate sizes) on the same Kenya + HealthBench
% tracks. Slot the table here and feed the result back into finding 3 in the
% intro. Until then, finding 3 leans on the 3x3 matrix + faithfulness only.
