\section{Retrieval Evaluation}
\label{sec:retrieval-results}

Having seen that retrieval barely moves end-to-end quality
(\S\ref{sec:endtoend}), we now isolate the retriever to ask whether the
problem is \emph{what} is retrieved. We score it on mamaretrieval---3{,}185
clinical queries with graded ($0$--$6$) relevance labels over the deployment
corpus, judged by Qwen3.5-397B and validated against Claude~Opus~4.7
(construction in the companion paper)---at the deployed retrieval depth of
top-3.

Because all three retrieved chunks are injected into the prompt together, the
operationally meaningful questions are whether any useful chunk is present
(Hit Rate) and how much of the small top-3 bundle is useful rather than noise
(Precision); rank-sensitive metrics such as MRR or nDCG would matter only if
order within the top-3 strongly shaped the answer, which we do not assume. We
report Hit Rate and Precision at a lenient threshold (relevance $\ge 3$, any
useful content) and a strict one ($\ge 5$, complete/specific content),
counting queries with no qualifying chunk as Hit Rate $=$ Precision $= 0$, as
a user would experience them.

We also report threshold-free weighted variants (wHR, wP) that replace the
binary relevance test with each chunk's graded score normalised to $[0,1]$
(the $0$--$6$ grade divided by $6$, so a $3$ contributes $0.5$). For each
query, wHR is the highest normalised grade among its top-3 chunks and wP is
the mean normalised grade across the three slots (empty slots counting $0$);
both are averaged over queries---the graded analogues of Hit Rate (best chunk
present) and Precision (average bundle quality).

\subsection{On-device retrieval is strong}

Figure~\ref{fig:retrievers} scores seven retrievers: the deployed on-device
EmbeddingGemma-300M, the earlier on-device Gecko, three cloud embedders
(voyage-4-large, octen, lateon), and lexical/biomedical baselines (BM25,
MedCPT). EmbeddingGemma was added to the original six-retriever pool in a
matched run under the same judge; the six reproduce exactly, confirming no
drift (companion paper).

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\footnotesize]
  % horizontal precision bars; x = P*5.0; lenient (light) drawn over strict (dark)
  \fill[mamteal!8] (-2.85,4.06) rectangle (5.0,4.74); % deployed-row highlight
  \foreach \p in {0.2,0.4,0.6,0.8}{%
    \draw[black!12] (\p*5,0.30) -- (\p*5,6.70);%
    \node[black!50, font=\scriptsize, below] at (\p*5,0.27) {\p};%
  }
  \draw[black!45] (0,0.30) -- (0,6.70);
  \node[black!60, font=\scriptsize] at (2.2,-0.18)
    {Precision@3 (fraction of the top-3 bundle that is relevant)};
  % legend
  \fill[mamteal!35] (0.0,7.02) rectangle (0.30,7.24);
  \node[anchor=west, font=\scriptsize] at (0.36,7.13) {lenient ($\ge3$)};
  \fill[mamteal] (2.05,7.02) rectangle (2.35,7.24);
  \node[anchor=west, font=\scriptsize] at (2.41,7.13) {strict ($\ge5$)};
  % rows: name / label-color / x_lenient / x_strict / P_lenient / P_strict / y
  \foreach \nm/\col/\xl/\xs/\vl/\vs/\y in {%
    {voyage}/black/4.335/2.260/0.867/0.452/6.30,%
    {octen}/black/4.020/2.015/0.804/0.403/5.35,%
    {{\bfseries EmbeddingGemma}}/mamteal/3.920/1.940/0.784/0.388/4.40,%
    {lateon}/black/3.690/1.750/0.738/0.350/3.45,%
    {Gecko}/mamteal/2.385/0.965/0.477/0.193/2.50,%
    {BM25}/black!55/2.085/0.815/0.417/0.163/1.55,%
    {MedCPT}/black!55/1.670/0.560/0.334/0.112/0.60}{%
    \node[anchor=east, text=\col] at (-0.12,\y) {\nm};%
    \fill[mamteal!35] (0,\y+0.04) rectangle (\xl,\y+0.30);%
    \fill[mamteal] (0,\y-0.30) rectangle (\xs,\y-0.04);%
    \node[anchor=west, font=\scriptsize] at (\xl+0.06,\y+0.17) {\vl};%
    \node[anchor=west, font=\scriptsize, text=mamteal] at (\xs+0.06,\y-0.17) {\vs};%
  }
\end{tikzpicture}
\caption{Retriever scoreboard on mamaretrieval ($n=3{,}185$, top-3), as top-3
precision at the lenient ($\ge3$, light) and strict ($\ge5$, dark) relevance
thresholds; longer is better. The deployed on-device EmbeddingGemma
(highlighted) lands third of seven, between cloud embedders and within
$\sim$8 points of the best (voyage), and far above the former on-device
Gecko. On-device retrievers are labelled in teal, cloud in black,
lexical/biomedical baselines in grey. Hit Rate and the threshold-free
weighted variants (wHR, wP) track the same ordering (see text).}
\label{fig:retrievers}
\end{figure}

The deployed retriever lands in the top tier---third of seven, between two
cloud embedders, within $\sim$8 points of precision of the best (voyage). A
300M model running on the phone competes with cloud retrieval, and it
improves on the earlier on-device Gecko by $+30.7$ points of lenient
precision (0.784 vs.\ 0.477), closing essentially the whole on-device gap.
The threshold-free weighted precision tells the same story (EmbeddingGemma
$\text{wP}=0.619$ vs.\ Gecko $0.393$, voyage $0.682$), as do both Hit-Rate
variants. One honest caveat is in Hit Rate at the strict bar, not in
precision: even the best retriever carries a complete, specific ($\ge5$)
chunk in its top-3 for only $\sim$75\% of queries (voyage $\text{HR}=0.753$;
deployed EmbeddingGemma $0.704$). At the deployed depth, then, a meaningful
minority of queries have no fully specific chunk in the bundle---a gap that
deeper retrieval or a broader corpus, rather than a better top-3 embedder,
would have to close.

\subsection{Better retrieval does not reach the answers}
\label{sec:retr-converts}

If retrieval quality were the binding constraint, improving it should lift
end-to-end answers. Two independent interventions say otherwise---upgrading
the embedder (Gecko$\to$EmbeddingGemma) and adding a fine-tuned reranker. Each
sharply improves offline retrieval, yet neither moves the answers: across
both, retrieval precision rises while end-to-end Kenya key-fact recall and
HealthBench stay flat (Figure~\ref{fig:retr-converts}).

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{groupplot}[
  group style={group size=1 by 2, vertical sep=1.4cm},
  width=\columnwidth, height=3.9cm,
  tick label style={font=\footnotesize}, label style={font=\footnotesize},
  grid=both, grid style={black!8},
  title style={font=\footnotesize\bfseries, yshift=-1ex},
]
% --- Panel a: offline retrieval (x = metric, bars = the 4 retrieval setups) ---
\nextgroupplot[ybar, bar width=6pt, area legend,
  title={Offline retrieval quality}, ylabel={precision@3},
  symbolic x coords={bench, kenya}, xtick=data,
  xticklabels={mamaretrieval P@3, Kenya P@3},
  enlarge x limits=0.5, ymin=0, ymax=0.88, ytick={0,0.2,0.4,0.6,0.8}]
  \addplot[fill=mamamber!85, draw=mamamber!70] coordinates {(bench,0.485)(kenya,0.270)}; \label{s:gecko}
  \addplot[fill=mamteal!80, draw=mamteal!60] coordinates {(bench,0.784)(kenya,0.396)}; \label{s:eg}
  \addplot[fill=blue!55!black, draw=blue!45!black] coordinates {(bench,0.720)(kenya,0.244)}; \label{s:minilm}
  \addplot[fill=violet!72, draw=violet!55] coordinates {(bench,0.794)(kenya,0.279)}; \label{s:mxbai}
% --- Panel b: end-to-end answers (x = metric, bars = the 5 setups incl no-RAG) ---
\nextgroupplot[ybar, bar width=3pt, area legend,
  title={End-to-end answers \quad (baseline $\,\vert\,$ deployed G1)}, ylabel={answer score},
  symbolic x coords={recall, wmet, pos, pen, gap, recallG, wmetG, posG, penG}, xtick=data,
  xticklabels={key-fact recall, weighted\_met, Positive, Penalty, {}, key-fact recall, weighted\_met, Positive, Penalty},
  x tick label style={rotate=30, anchor=east, font=\scriptsize},
  enlarge x limits=0.04, ymin=-0.04, ymax=0.42, ytick={0,0.1,0.2,0.3,0.4}]
  \draw[densely dotted, black!60, thick] (axis cs:gap,-0.04) -- (axis cs:gap,0.42);
  % left block = baseline prompt (5 setups); right block = deployed G1 (3 setups)
  \addplot[fill=black!42, draw=black!50] coordinates {(recall,0.178)(wmet,0.003)(pos,0.184)(pen,0.379)(recallG,0.328)(wmetG,0.054)(posG,0.229)(penG,0.371)}; \label{s:norag}
  \addplot[fill=mamamber!85, draw=mamamber!70] coordinates {(recall,0.125)(wmet,-0.004)(pos,0.175)(pen,0.372)(recallG,0.268)(wmetG,0.029)(posG,0.216)(penG,0.390)};
  \addplot[fill=mamteal!80, draw=mamteal!60] coordinates {(recall,0.126)(wmet,0.012)(pos,0.189)(pen,0.368)(recallG,0.279)(wmetG,0.038)(posG,0.221)(penG,0.373)};
  \addplot[fill=blue!55!black, draw=blue!45!black] coordinates {(recall,0.118)(wmet,-0.004)(pos,0.179)(pen,0.378)};
  \addplot[fill=violet!72, draw=violet!55] coordinates {(recall,0.125)(wmet,0.002)(pos,0.182)(pen,0.374)};
\end{groupplot}
% legend: the five retrieval setups (single box, one row, single swatches)
\node[anchor=north, font=\scriptsize, inner sep=0pt, yshift=-0.3cm]
  at (current bounding box.south) {%
  \setlength{\fboxsep}{4pt}\setlength{\fboxrule}{0.4pt}%
  \fcolorbox{black!30}{white}{\textbf{Setup:}\enspace\ref{s:norag}~no-RAG\enspace\ref{s:gecko}~Gecko\enspace\ref{s:eg}~EmbeddingGemma\enspace\ref{s:minilm}~+MiniLM\enspace\ref{s:mxbai}~+mxbai}%
};
\end{tikzpicture}
\caption{Better retrieval does not reach the answers, and a stronger prompt
does not change that. Each x-group is one metric and the bars are the
retrieval setups, so setups compare directly within a metric.
\textbf{Top, offline retrieval (baseline Gemma~4):} on the mamaretrieval
benchmark, EmbeddingGemma and both fine-tuned rerankers (on the Gecko first
stage) lift P@3 from Gecko's 0.49 to 0.72--0.79; on the Kenya queries only the
embedder improves. \textbf{Bottom, end-to-end answers:} within every
metric---key-fact recall and the HealthBench net \texttt{weighted\_met} with
its completeness (Positive) and penalty components---the setups are
essentially level, so retrieval does not move the answers (no-RAG leads on
recall). \emph{Left of the dotted line} is the baseline prompt; \emph{right}
is the deployed Gemma~4\,+\,G1 prompt (no-RAG/Gecko/EmbeddingGemma only;
rerankers were run under baseline alone). G1 lifts every answer metric
(deflection $33\%\!\to\!3\%$) yet the ranking is unchanged---no-RAG still
leads and RAG stays net-negative. Net \texttt{weighted\_met} $\approx0$ is
separately normalized, not Positive minus Penalty; EmbeddingGemma (teal) is
deployed.}
\label{fig:retr-converts}
\end{figure}

The embedder bake-off is the cleaner of the two. Gecko$\to$EmbeddingGemma
raises Kenya retrieval precision $0.270\to0.396$ ($+12.6$ points), but
end-to-end key-fact recall stays flat ($0.125\to0.126$) and HealthBench barely
moves. More starkly, \emph{no retrieval at all} scores highest on Kenya
(recall 0.178): on this small generator, adding context---however good---does
not help, and slightly hurts.

The reranker makes the same point from a second angle. It is a
\emph{cross-encoder} second stage that re-scores the first stage's top-20
(query, chunk) pairs jointly---more accurate than the bi-encoder retriever but
too costly to run corpus-wide. We screened several deployable cross-encoders
against multi-billion-parameter Qwen3-Reranker references and fine-tuned two
of them in-domain on 158{,}993 judge-graded (query, chunk) pairs
(\emph{fine-tuning, not size, was the lever}: the 23M MiniLM-L6 beats an
$8\times$-larger zero-shot model and ships int8 on-device). Both lift
\emph{benchmark} precision sharply, from 0.485 to 0.72--0.79---yet the gain
does not transfer: on the Kenya queries retrieval precision moves only to
0.24--0.28 and key-fact recall stays at 0.12 (net HealthBench $\approx0$). The lesson is methodological: retriever
changes must be validated on the deployment queries, not the benchmark alone,
and we ship no reranker.

Why do the retrieval gains not convert? The information is not missing:
precision rises, so it is retrieved; the corpus holds it
(\S\ref{sec:retr-corpus}); and it cannot be gated away
(\S\ref{sec:retr-gate}). The break is downstream, in the small generator's
response behaviour---which the prompt matrix (\S\ref{sec:endtoend}) already
exposed and better context does not change. It declines to engage, or engages
without committing to specific guidance (``see a doctor''); for such answers,
no improvement in the retrieved chunk can raise key-fact recall. (A subtler,
metric-level contributor: retrieval precision rewards \emph{topical}
relevance, which need not coincide with the specific gold facts key-fact
recall scores.)

Nor is the baseline prompt's 33\% deflection masking the gain: re-running the
embedder bake-off under the deployed Gemma~4\,+\,G1 prompt, which cuts
deflection to 3\%, leaves the ranking unchanged (Figure~\ref{fig:retr-converts},
right of the dotted line). No-RAG still leads (recall 0.328), EmbeddingGemma
(0.279) again edges Gecko (0.268) but neither reaches it, and RAG stays
net-negative (the harm rate is likewise flat at $\sim$0.16)---an effect the
size ladder (\S\ref{sec:model-size}) shows deepening with capability (the RAG
penalty grows from $-0.01$ HealthBench at 4B to $-0.10$ at 31B). The binding
constraint is the generator's use of context, which we isolate next
(\S\ref{sec:generator}).

\subsection{The corpus is mostly, not wholly, sufficient}
\label{sec:retr-corpus}

Nor is the corpus the \emph{primary} gap. A coverage audit pooling the top-20
of all seven retrievers---their union, so a test of corpus reachability rather
than of any single retriever---finds a relevant chunk present for $96.5\%$ of
Kenya queries at the lenient bar and $89.4\%$ at the strict bar; only $10.6\%$
are genuinely absent from the corpus. For most queries, then, a complete
answer is both in the corpus and retrievable, and the break is at the last
step---the generator's use of what it is handed. This is the prompt- and
retriever-independent counterpart to the bake-off and the faithfulness
analysis (\S\ref{sec:generator}): the binding constraint is the generator.

Two corpus limits nonetheless remain, as genuine but secondary constraints.
First, the $10.6\%$ of queries with no relevant chunk anywhere are unreachable
by any retriever or generator---a hard content gap that only corpus expansion
can close. Second, even when a relevant chunk is present its quality is often
modest: the deployed retriever's top-3 carries a mean relevance grade of just
$2.13/6$ (Gecko $1.44/6$), so the corpus tends to supply topical rather than
complete-and-specific support for these nurse-written vignettes. An earlier
reading took this low absolute relevance (mean grade $1.49/6$) as evidence
that the corpus was itself the deeper bottleneck; the union audit reframes
it---the content is mostly reachable at depth-20, so the low top-3 grade
reflects ranking and depth, not absence. Corpus expansion therefore stays a
worthwhile but lower-priority lever (\S\ref{sec:discussion}), behind closing
the generator's grounding gap.

\subsection{No usable confidence gate}
\label{sec:retr-gate}

A natural mitigation for ``RAG sometimes hurts'' is an abstention gate:
detect a weak retrieval from its score and fall back to no-RAG. We re-asked
this for the deployed EmbeddingGemma---an earlier study had shelved it for
Gecko---and the verdict is the same: the retrieval score is not thresholdable
into a usable gate. Every relevance signal a gate could key on falls below the
standard $0.80$ AUC viability bar (Figure~\ref{fig:gate-auc}). Cosine
separates relevant from irrelevant Kenya chunks at AUC only $0.72$, and the
three threshold-rule families a gate could implement---an absolute cutoff on
chunk relevance, the top-1 chunk's margin over the rest of the bundle, and
gated abstention on whether the bundle holds any relevant chunk---sit at
$0.69$--$0.72$, with the relative-margin rule near chance. These are pure
retrieval-relevance statistics, computed on EmbeddingGemma's top-3 against
Qwen-judged chunk relevance, so they are independent of the generator and
prompt and hold for the deployed configuration.

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\footnotesize]
  % AUC axis 0.50..0.85 mapped to x=(v-0.50)*18: chance at x=0, 0.80 bar at x=5.4
  \draw[mamamber, dashed, thick] (5.4,0) -- (5.4,3.35);
  \node[mamamber, font=\scriptsize, above] at (5.4,3.35) {$0.80$ viability bar};
  \draw[black!45] (0,0) -- (0,3.35);
  \node[black!55, font=\scriptsize, below] at (0,0) {$0.50$ (chance)};
  \foreach \nm/\xv/\val/\y in {%
    {Chunk separability (pooled AUC)}/3.96/0.720/3.0,%
    {Gated abstention (bundle top-1 AUC)}/3.888/0.716/2.1,%
    {Absolute cutoff (top-3 chunk AUC)}/3.456/0.692/1.2,%
    {Relative margin (within-query concord.)}/1.188/0.566/0.3}{%
    \node[anchor=west, font=\scriptsize] at (0,\y+0.30) {\nm};%
    \fill[mamteal!78] (0,\y-0.15) rectangle (\xv,\y+0.15);%
    \node[anchor=west, font=\scriptsize] at (\xv+0.08,\y) {$\val$};%
  }
\end{tikzpicture}
\caption{No retrieval-relevance signal is thresholdable into a usable gate.
Each bar is the AUC of an EmbeddingGemma cosine statistic at separating
relevant ($\ge3$) from irrelevant Kenya chunks ($n=6{,}228$ judged top-20
pairs); all four fall below the standard $0.80$ viability bar, and the
relative-margin rule is near chance. These are pure retrieval-relevance
metrics, independent of the generator and prompt; the same four statistics
left the earlier Gecko retriever ungateable too.}
\label{fig:gate-auc}
\end{figure}

Zooming into the cutoff rule, no operating point is usable either: against a
$\sim$20\% relevant base rate, tightening the cosine cutoff lifts
strict-relevant precision only to $0.41$ while recall collapses
(Table~\ref{tab:gate}).

\begin{table}[t]
\caption{Why a cosine relevance gate fails at the operating-point level. As
the cutoff on EmbeddingGemma's similarity score is tightened, strict-relevant
($\ge5$) precision rises only to $0.41$ while recall collapses---no operating
point has both, against a $\sim$20\% relevant base rate.}
\label{tab:gate}
\centering
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
Cosine cutoff & Chunks kept & \makecell{Precision\\(strict) $\uparrow$} & \makecell{Recall\\(strict) $\uparrow$} \\
\midrule
$\ge 0.50$ & 70\% & 0.260 & 0.900 \\
$\ge 0.55$ & 36\% & 0.345 & 0.627 \\
$\ge 0.60$ & 11\% & 0.413 & 0.235 \\
\bottomrule
\end{tabular}
\end{table}

A separate, stronger test asks not whether the score finds relevant chunks but
whether it predicts when RAG actually \emph{helps the answer}: it does not---%
AUC $0.50$, Pearson $-0.04$, pure chance---and a perfect oracle gate would lift
recall only $0.297\to0.354$. This answer-side test, however, was run on
Gemma~3n under the baseline prompt ($n=300$), not the deployed
Gemma~4\,+\,G1, so---like the bake-off (\S\ref{sec:retr-converts})---it awaits
confirmation under the shipped configuration; the relevance-side verdict above
does not depend on it. Either way, a usable confidence signal would have to
come from the generator (answer-side confidence or self-consistency), not from
the retrieval score.
% TODO (with the G1 bake-off, sec:retr-converts): re-run the answer-side gate test
% (score -> "RAG helps") on the deployed Gemma 4 + G1, not the 06-18
% value_gate_eg_3n run (Gemma 3n + baseline). The relevance-side AUCs
% (fig:gate-auc) are prompt/generator-independent and need no re-run.

\medskip
Every retrieval lever points the same way. The deployed retriever is strong
and competes with the cloud; a better embedder and a fine-tuned reranker both
win offline; the corpus covers most questions; and no confidence gate can
sidestep the bad cases---yet end-to-end answers do not move. The binding
constraint is the small on-device generator's ability to turn retrieved
context into better answers, which we examine directly next
(\S\ref{sec:generator}) and whose remedy---grounding the generator---we take
up in the discussion (\S\ref{sec:discussion}).
