\section{Generator Faithfulness}
\label{sec:generator}

The end-to-end matrix (\S\ref{sec:helpful-safe}) chose Gemma~4 over the more
helpful Gemma~3n on a single decisive signal: the count of \emph{dangerous}
Kenya answers ($0$/$1$/$0$ vs.\ $4$/$11$/$15$). That count is small and
benchmark-specific, so a deployment decision should not rest on it alone. This
section supplies an independent, second line of evidence from a different
angle---how faithfully each generator stays inside its sources---and it
returns the same verdict. We isolate the generator by replacing live retrieval
with \emph{oracle} (gold) context, so any failure is the generator's own
grounding behaviour, not a retrieval miss.

\subsection{Setup: faithfulness to gold context}
\label{sec:faith-setup}

Each generator is handed the gold guideline passages for the same 2{,}989
maternal/neonatal questions and asked to answer. The context is the
mamaretrieval passages graded $\geq 5$ (the relevance ceiling),
top-3---matched to the deployed retrieval depth---and \emph{identical} across
every generator and prompt, so any difference in the results is attributable
to the generator or the prompt, never to retrieval. We run the same
$3\times3$ matrix as \S\ref{sec:endtoend}: the two on-device candidates
(Gemma~4~E4B and Gemma~3n~E4B, both at the on-device int4/Q4\_0 quantization)
and the Qwen3.5-397B frontier ceiling, each under baseline, \mbox{+G1}, and
\mbox{+G1+G2}.

\emph{Faithfulness} asks one thing: does the answer stay inside what the
context supports? We count two failure modes as genuine unfaithfulness:
\begin{itemize}\setlength{\itemsep}{1pt}
  \item \textbf{Contradiction} --- the answer asserts something the context
  denies (a wrong dose, a wrong threshold, states X where the source says Y).
  \item \textbf{Unsupported addition} --- the answer makes a clinical claim
  the context does not support (information introduced beyond the document).
\end{itemize}
Incompleteness (\emph{omission}) and \emph{refusal} are deliberately
\emph{not} faithfulness failures: an answer that leaves something out or
declines to engage is unhelpful, but it is not \emph{unfaithful}---it invents
nothing. Those behaviours are scored on the helpfulness side
(\S\ref{sec:endtoend}); here they would only inflate the rate. Because the
context is gold, this measures grounding in isolation and is an \emph{upper
bound}: under live retrieval, missing or wrong passages add a separate error
mode (\S\ref{sec:retrieval-results}) not captured here.

\subsection{A two-pass judge}
\label{sec:faith-judge}

The rubric judge of \S\ref{sec:judge} (gpt-oss-120b) is not used here:
detecting hallucination is a different task from scoring a rubric, and
gpt-oss-120b in fact failed the pre-committed validation gates for it,
drifting badly---it labelled refusals and omissions as hallucinations, exactly
the confound we must avoid. We therefore use a dedicated two-pass pipeline,
a high-recall detector followed by a precise categorizer:
\begin{itemize}\setlength{\itemsep}{1pt}
  \item \textbf{Detector --- Patronus Lynx-70B.} A purpose-built hallucination
  detector returns PASS/FAIL with reasoning for every answer. Lynx is
  high-recall but low-precision (precision $0.36$--$0.57$ across arms): it
  catches almost everything but over-flags, marking incompleteness and
  refusals as failures.
  \item \textbf{Categorizer --- gpt-5.} Every Lynx FAIL is re-read against a
  frozen rubric and sorted into contradiction / unsupported addition /
  omission / refusal / unclear; only the first two count. This strips Lynx's
  over-flags and keeps the genuine hallucinations.
\end{itemize}
The metric of record is the \textbf{categorized true-hallucination
rate}---genuine failures as a fraction of all 2{,}989 answers, computed over
the full population with no extrapolation, hence deterministic and
reproducible. gpt-5 is a different model family from Lynx (Llama-3), from the
generators (Gemma), and from the oracle's relevance labeler (Qwen), so the
audit shares no blind spot with what it audits.

\subsection{Gemma 4 grounds about as faithfully as the frontier}
\label{sec:faith-results}

Figure~\ref{fig:faithfulness} is the headline. Two facts hold across every
prompt:

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\columnwidth, height=5.4cm,
  ybar, bar width=7.5pt,
  ymin=0, ymax=8,
  ylabel={Categorized true-hallucination},
  ylabel style={align=center},
  yticklabel={\pgfmathprintnumber{\tick}\%},
  ytick={0,2,4,6,8},
  symbolic x coords={base,g1,g2},
  xtick=data,
  xticklabels={baseline,+G1,+G1+G2},
  enlarge x limits=0.32,
  tick label style={font=\footnotesize},
  label style={font=\footnotesize},
  legend style={font=\footnotesize, at={(0.5,-0.20)}, anchor=north,
    legend columns=3, draw=black!30, /tikz/every even column/.append style={column sep=6pt}},
  area legend,
  nodes near coords, every node near coord/.append style={font=\tiny,
    /pgf/number format/fixed, /pgf/number format/fixed zerofill,
    /pgf/number format/precision=2},
]
\addplot[fill=mamteal, draw=mamteal!65!black] coordinates {(base,2.64)(g1,3.31)(g2,3.55)};
\addplot[fill=mamamber, draw=mamamber!65!black] coordinates {(base,6.26)(g1,6.46)(g2,6.72)};
\addplot[fill=black!22, draw=black!55] coordinates {(base,2.51)(g1,3.71)(g2,0.97)};
\legend{Gemma 4 E4B (deployed), Gemma 3n E4B, Qwen 397B (frontier)}
\end{axis}
\end{tikzpicture}
\caption{Categorized true-hallucination rate against gold context (lower is
better), over $n=2{,}989$ oracle answers per cell, two-pass Lynx-70B$\,\to\,$
gpt-5 judge. Gemma~3n hallucinates against the very context it is handed
roughly twice as often as Gemma~4 at every prompt, while Gemma~4 grounds about
as faithfully as the unconstrained 397B frontier. The prompt moves each model
only slightly; the gap between models dwarfs it.}
\label{fig:faithfulness}
\end{figure}

\begin{itemize}\setlength{\itemsep}{2pt}
  \item \textbf{Gemma~4 is $\sim$2$\times$ more faithful than Gemma~3n.} On
  the categorized rate, Gemma~4 sits at $2.6$--$3.6\%$ across the three
  prompts while Gemma~3n sits at $6.3$--$6.7\%$---a $1.9$--$2.4\times$ gap that
  holds for baseline, \mbox{+G1}, and \mbox{+G1+G2} alike.
  \item \textbf{Small size is not the problem; Gemma~3n is the outlier.}
  Gemma~4, a 4B model running int4 on a phone, grounds about as faithfully as
  the 397B Qwen frontier ($2.5$--$3.7\%$). Faithfulness is therefore not a
  penalty of the on-device size class---it is a property of the specific
  model, and Gemma~3n is the one that grounds poorly.
\end{itemize}

The prompt's effect is real but small: both models drift \emph{upward} as the
prompt asks for more clinical substance (Gemma~4 by $+0.9$ percentage points
from baseline to \mbox{+G1+G2}, Gemma~3n by $+0.5$), because the recovered
substance sometimes reaches beyond the document. But that drift is dwarfed by
the gap between the two models---choosing the on-device generator matters
roughly $4\times$ more for faithfulness than choosing the prompt.

\subsection{What kind of unfaithfulness, and how the prompt changes it}
\label{sec:faith-modes}

Splitting each Lynx FAIL by primary cause (Table~\ref{tab:faith-modes}) shows
\emph{why} the rates move and ties the faithfulness cost back to the
deflection fix of \S\ref{sec:endtoend}.

\begin{table}[t]
\caption{Failure-cause decomposition of the oracle answers (counts out of
$n=2{,}989$; Qwen omitted as the frontier reference). The two genuine
hallucination modes---contradiction and unsupported addition---sum to the
categorized rate in the last column (Figure~\ref{fig:faithfulness}); omission
and refusal are over-flags that do \emph{not} count. Reading the rows: \mbox{+G1}
trades refusals for unsupported additions (Gemma~4 refusal $27\to7$ while
unsupported additions $28\to55$), and Gemma~3n carries roughly twice Gemma~4's
contradiction load at every prompt.}
\label{tab:faith-modes}
\centering
\small
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}lrrrrr@{}}
\toprule
 & \multicolumn{2}{c}{Genuine ($\downarrow$)} & \multicolumn{2}{c}{Over-flags (not counted)} & \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Generator $\cdot$ prompt & Contradiction
 & \makecell{Unsupported\\addition} & Omission & Refusal
 & \makecell{Categorized\\rate} \\
\midrule
\multicolumn{6}{@{}l}{\textit{Gemma 4 E4B (deployed)}} \\
\quad baseline & 51 & 28 & 63 & 27 & 2.64\% \\
\quad +G1      & 44 & 55 & 26 & \phantom{0}7 & 3.31\% \\
\quad +G1+G2   & 48 & 58 & 36 & 20 & 3.55\% \\
\midrule
\multicolumn{6}{@{}l}{\textit{Gemma 3n E4B}} \\
\quad baseline & 91 & \phantom{0}96 & 47 & 14 & 6.26\% \\
\quad +G1      & 88 & 105 & 30 & 103 & 6.46\% \\
\quad +G1+G2   & 88 & 113 & 44 & 70 & 6.72\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{itemize}\setlength{\itemsep}{2pt}
  \item \textbf{G1 converts refusals into unsupported additions.} For
  Gemma~4, refusal failures fall $27\to7$ while unsupported additions rise
  $28\to55$: the prompt stops the model deflecting, and the substance it now
  volunteers sometimes reaches beyond the document. This is the deflection fix
  of \S\ref{sec:endtoend} working \emph{and} its price, in a single row---the
  faithfulness counterweight to the recall gain that motivated G1.
  \item \textbf{Gemma~3n's deficit is a baseline contradiction load.} Before
  any prompt change, Gemma~3n contradicts its context almost twice as often as
  Gemma~4 ($91$ vs.\ $51$): it gets clinical facts wrong against the very
  passages it was handed. This is the core grounding gap, and it is exactly
  what surfaces end-to-end as Gemma~3n's dangerous Kenya answers
  (\S\ref{sec:helpful-safe})---poor grounding becomes confident, specific,
  unsupported guidance, which becomes harm.
\end{itemize}

\subsection{An honesty note on the calibrated estimate}
\label{sec:faith-calibration}

Lynx is high-recall but not perfect---it also \emph{misses} some true
failures. A second, \emph{calibrated} estimate corrects for those misses by
re-judging a 50-answer sample of Lynx-PASS cases per arm with gpt-5 and
extrapolating the miss rate across the $\sim$2{,}800 PASS answers. Because it
multiplies a 50-sample rate across the whole population, it is noisy, with wide
confidence intervals---we report the categorized rate, not this one. One
calibrated number is worth flagging precisely because it looks alarming out of
context: Gemma~4~+G1+G2 calibrates to $53.9\%$, far above its neighbours. This
is a measurement artifact---an effect of how the rate is estimated---not a real
drop in how faithfully the model grounds its answers. Three things stack up.
The G1+G2 consultation template makes answers $\sim$60\% longer (median
$1{,}193$ vs.\ $742$ characters); the calibration rubric judges strictly
against the \emph{narrow} top-3 context only, so clinically-sound elaboration
that is correct but simply absent from those three passages is counted as an
``unsupported addition''; and the estimate then scales one unusually strict
50-answer sample across the whole $\sim$2{,}800-answer population. The number
therefore reflects long answers exceeding a deliberately narrow context, not
Gemma~4 inventing false claims at scale: the full-population
\emph{categorized} rate for that same cell is $3.55\%$---still second-lowest in
the matrix. The calibrated column does corroborate the two qualitative facts
above (Gemma~3n $>$ Gemma~4 at every arm; prompts push the rate up), but only
with wider error bars; its magnitude is not reliable, and we do not lean on it.

\subsection{Implication: faithfulness is a generator property}
\label{sec:faith-implication}

Two independent angles now converge on the same conclusion. The Kenya
dangerous count (\S\ref{sec:helpful-safe}) and oracle faithfulness (this
section) are measured on different data, with different judges, against
different references---yet both find Gemma~4 roughly $2\times$ safer than
Gemma~3n, and both locate the cause in the model rather than the prompt. This
is the deciding evidence behind deploying Gemma~4: it is the candidate that is
as faithful as a frontier model while remaining small enough to run on the
phone.

The deficit is structural. No prompt in this study closes Gemma~3n's gap;
\mbox{+G1}/\mbox{+G2} modestly widen it, and the gap is roughly $4\times$ the
prompt effect. So the path to a more faithful---and therefore safely more
helpful---on-device model is generator-side work: RAG-grounding or fine-tuning
the generator to use only what it retrieves, not further prompt engineering or
retrieval swaps. This is the same lever the size ladder
(\S\ref{sec:model-size}) and the retrieval analysis
(\S\ref{sec:retrieval-results}) point to, and we take it up in
\S\ref{sec:discussion}. Two caveats bound the claim: this is faithfulness to
\emph{gold} context (an upper bound; live retrieval adds its own errors), and
the semantic categorization rests on a single judge family
(\S\ref{sec:limitations}).
