\section{On-Device Latency}
\label{sec:latency}

The preceding sections establish that the system answers \emph{well}; this one
establishes that it answers \emph{fast enough}, on the device, with no network.
The app ships \textbf{CPU-by-default} for broad compatibility---the safer,
lowest-common-denominator backend---with GPU execution as an opt-in that has a
higher hardware bar (\S\ref{sec:devices}). At the deployed top-3 depth, Gemma~4
E4B returns a full answer in a median of \textbf{$\sim$43\,s} on the shipped CPU
path and \textbf{$\sim$19\,s} where a compatible GPU is available (first token
in $\sim$1\,s, generation at $\sim$13~tokens/s). Both fit the field budget we
set ourselves (a query should resolve well under a minute) on this flagship and
run entirely offline---the GPU with wide headroom, the CPU with a tighter margin
that shrinks further on the cheaper hardware typical of the field
(\S\ref{sec:devices}). The
rest of this section traces where the time goes, surfaces one precision bug that
bounds how much context we can admit, and sets out what a device needs to run it
at all---the binding deployment constraint, it turns out, is not speed but
memory and GPU availability.

\subsection{Setup}
\label{sec:latency-setup}

All numbers are measured on real hardware: an OnePlus tablet (Snapdragon~8
Elite, 16\,GB RAM, Android~15) running LiteRT-LM 0.11.0. We sweep three axes
that matter for deployment---generator size (Gemma~4 E4B vs the smaller E2B),
compute backend (GPU via OpenCL on the Adreno; CPU via XNNPACK), and retrieval
depth $k\in\{0,1,3,5,7,10,15,20\}$---and time every cell over 18 query types
$\times$ 3 repeats ($54$ runs), with a 10\,s cooldown between runs for thermal
stability. We separate three quantities: \emph{time-to-first-token} (TTFT, the
prefill pass over the prompt, excluding retrieval), \emph{decode} (first to
last generated token), and \emph{total query} (retrieval $+$ TTFT $+$ decode,
what the user waits). Each benchmark records the SHA-256 fingerprint of the
loaded model artifact so a reviewer can verify which weights produced which
timing.

Retrieval---embedding the query and searching the vector store---is timed
separately from TTFT and adds a near-constant $\sim$1.7\,s at the deployed
depth, almost all of it the single query-embedding pass rather than the
similarity search over the $63{,}650$-chunk index (which is negligible beside
it). That figure predates the Gecko~$\to$~EmbeddingGemma swap
(\S\ref{sec:system-design}) and may have shifted, but retrieval is a small slice
next to the generation that dominates the total (the $\sim$16\,s decode at the
deployed $k{=}3$), and the deployed generator is unchanged---so the
generation-set headline figures below stand.

One further mismatch with the deployed configuration, also minor: these runs
predate the G1 prompt (\S\ref{sec:system-prompt}) and use the earlier baseline
prompt, which deflected on about a third of questions with short ``see a
doctor'' replies. Because G1 cuts deflection to $\sim$3\%, the deployed system
produces longer, more substantive answers on average, so its decode---and thus
total latency---will run somewhat higher than measured here. Like the retriever
change, this leaves the picture generation-dominated and within budget: the
figures are a mild \emph{under}-estimate of the deployed system, not an
over-estimate.

\subsection{Latency by backend and retrieval depth}
\label{sec:latency-sweep}

Figure~\ref{fig:latency-k} is the headline. Three things stand out.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\columnwidth, height=5.6cm,
  xlabel={Retrieval depth $k$ (passages in context)},
  ylabel={Median total query (s)},
  xmin=-0.6, xmax=15.6, ymin=0, ymax=92,
  xtick={0,1,3,5,7,10,15}, xticklabels={0,1,3,5,7,10,15},
  ytick={0,20,40,60,80},
  grid=both, grid style={black!8},
  tick label style={font=\footnotesize}, label style={font=\footnotesize},
  legend style={font=\footnotesize, at={(0.5,-0.26)}, anchor=north,
    legend columns=2, draw=black!30, /tikz/every even column/.append style={column sep=10pt}},
]
  % 60s field budget
  \draw[red!55, dashed, thick] (axis cs:-0.6,60) -- (axis cs:15.6,60);
  \node[font=\scriptsize, text=red!60, anchor=south east] at (axis cs:15.6,60) {60\,s field budget};
  % deployed retrieval-depth marker (top-3; backend-agnostic)
  \draw[black!35, dotted, thick] (axis cs:3,0) -- (axis cs:3,92);
  \node[font=\scriptsize, text=black!55, rotate=90, anchor=south] at (axis cs:3,46) {deployed ($k{=}3$)};
  \addplot[mamteal, very thick, mark=*, mark size=1.7pt] coordinates {
    (0,14.4)(1,14.1)(3,19.1)(5,19.6)(7,22.9)(10,22.4)(15,24.4)};
  \addplot[mamteal, very thick, densely dashed, mark=square*, mark size=1.5pt] coordinates {
    (0,28.0)(1,30.3)(3,42.7)(5,60.2)(7,62.3)(10,69.4)(15,84.9)};
  \addplot[mamamber, very thick, mark=triangle*, mark size=2pt] coordinates {
    (0,8.7)(1,11.7)(3,14.3)(5,11.6)(7,15.2)(10,15.6)(15,13.1)};
  \addplot[mamamber, very thick, densely dashed, mark=diamond*, mark size=1.8pt] coordinates {
    (0,13.9)(1,15.8)(3,20.6)(5,27.2)(7,28.5)(10,26.3)(15,36.8)};
  \legend{E4B GPU (opt-in), E4B CPU (shipped), E2B GPU, E2B CPU}
\end{axis}
\end{tikzpicture}
\caption{Median total query latency vs.\ retrieval depth on the Snapdragon~8
Elite, by generator and backend (lower is better; $k=20$ omitted---see
\S\ref{sec:context-wall}). The optional E4B-GPU path (solid teal) stays
$14$--$24$\,s across all depths, far under the 60\,s field budget; the shipped
E4B-CPU default (dashed teal) is slower and crosses the budget past
$k{\approx}5$, while the smaller E2B (amber) keeps every depth comfortably
inside it on either backend.}
\label{fig:latency-k}
\end{figure}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\columnwidth, height=5.2cm,
  xlabel={Retrieval depth $k$ (passages in context)},
  ylabel={Time (s), Gemma~4 E4B},
  xmin=-0.6, xmax=15.6, ymin=0, ymax=60,
  xtick={0,1,3,5,7,10,15}, xticklabels={0,1,3,5,7,10,15},
  ytick={0,20,40,60},
  grid=both, grid style={black!8},
  tick label style={font=\footnotesize}, label style={font=\footnotesize},
  legend style={font=\footnotesize, at={(0.5,-0.30)}, anchor=north,
    legend columns=2, draw=black!30, /tikz/every even column/.append style={column sep=10pt}},
]
  % CPU first-token (prefill) -- the one that explodes
  \addplot[mamamber, very thick, densely dashed, mark=square, mark size=1.3pt] coordinates {
    (0,12.6)(1,12.6)(3,18.4)(5,36.4)(7,36.4)(10,40.0)(15,54.7)};
  % CPU generation (decode)
  \addplot[mamamber, very thick, mark=*, mark size=1.3pt] coordinates {
    (0,15.3)(1,14.0)(3,19.1)(5,21.6)(7,23.5)(10,21.7)(15,22.5)};
  % GPU first-token (prefill)
  \addplot[mamteal, very thick, densely dashed, mark=square, mark size=1.3pt] coordinates {
    (0,0.96)(1,0.95)(3,0.99)(5,1.88)(7,1.92)(10,2.52)(15,3.46)};
  % GPU generation (decode)
  \addplot[mamteal, very thick, mark=*, mark size=1.3pt] coordinates {
    (0,13.5)(1,11.4)(3,16.4)(5,15.9)(7,17.2)(10,18.1)(15,16.8)};
  \legend{CPU first-token, CPU generation, GPU first-token, GPU generation}
\end{axis}
\end{tikzpicture}
\caption{Where the time goes as retrieval depth grows: first-token (prefill,
dashed) and generation (decode, solid) latency for the deployed E4B generator,
GPU (teal) vs.\ CPU (amber). On the \textbf{GPU} both curves are nearly flat in
$k$---added context is almost free. On the \textbf{CPU} generation stays flat
($\sim$14--23\,s) but first-token climbs steeply, from $\sim$13\,s to
$\sim$55\,s, because prefill is compute-bound and the CPU has far less compute.
That prefill blow-up---not decode---is what pushes the CPU total past budget in
Figure~\ref{fig:latency-k}.}
\label{fig:latency-decomp}
\end{figure}

\paragraph{The GPU path has wide headroom.} Gemma~4~E4B on the GPU stays
between $14$ and $24$\,s of total latency across \emph{every} retrieval depth up
to $k=15$---roughly a third of the 60\,s budget at the deployed $k=3$
($\sim$19\,s). Latency barely grows with $k$ because, on the GPU, the cost is
dominated by \emph{decode}, not by the prompt
(Figure~\ref{fig:latency-decomp}, teal): TTFT rises only from $\sim$1\,s at
$k=3$ to $\sim$3.5\,s at $k=15$, while decode---a near-constant
$\sim$13--18\,s---sets the floor. Adding retrieved context is close to free on
the GPU.

\paragraph{CPU is the real fork in the road.} On the CPU---the shipped
default---the same E4B generator runs $2$--$3.5\times$ slower than on the GPU
($28$--$85$\,s over the same range) and crosses the 60\,s budget past
$k{\approx}5$. The gap is almost entirely
prefill (Figure~\ref{fig:latency-decomp}, amber dashed): CPU TTFT balloons from
$\sim$13\,s ($k=0$) to $\sim$55\,s ($k=15$), because prefill is compute-bound
and the CPU has far less of it, while CPU \emph{generation} stays flat like the
GPU's. This is what makes the GPU/CPU split, not the retrieval depth, the
decisive latency variable.

\paragraph{The smaller generator lowers the hardware floor.} The unlock
from halving the generator to E2B is \emph{memory}, not speed. At the deployed
$k=3$, even E4B on CPU ($\sim$43\,s on our flagship) fits the 60\,s budget, so a
missing GPU does not by itself rule out the deployed model. What rules it out is
RAM: E2B halves the runtime footprint ($\sim$3.3 to $\sim$1.7\,GB), dropping the
device RAM floor from 6\,GB to 4\,GB and so running on devices that cannot hold
E4B at all (\S\ref{sec:devices}). E2B is also faster---$\sim$2.3$\times$ on
prefill and $\sim$1.5$\times$ (GPU) to $\sim$2$\times$ (CPU) on total, the
decode gain smaller because GPU decode is memory-bandwidth- rather than
compute-bound---which buys latency headroom on the slower mid-tier CPUs where
E4B could exceed budget, though that hardware is not yet measured
(\S\ref{sec:latency-pending}).

\subsection{The FP16-GPU context wall}
\label{sec:context-wall}

We cap the model's total context at $4{,}096$ tokens. That cap is not a memory
or latency limit---it is a workaround for a precision bug we found, and it is
worth documenting because it will bite anyone deploying a LiteRT-LM-class
runtime at longer context.

On Android, the GPU text-decoder path defaults to FP16 activations (the CPU
path uses FP32); this is a runtime choice, not a knob in our code. Past a total
context of $\sim$5{,}000 tokens, FP16 GPU decode collapses into a degenerate
repetition loop (a stream of \texttt{*} characters) and the output silently
becomes garbage. The collapse is \emph{deterministic}, not a stochastic glitch:
the GPU backend decodes greedily, so a fixed prompt produces bit-for-bit
identical garbage on every run (we confirmed bit-exact reproduction across
repeated runs of the reference case), which rules out sampling noise and points
to the FP16 attention math itself---once the precision error pushes the
attention scores onto the \texttt{*} token, the model self-reinforces into the
loop. The same prompt decoded on the FP32 CPU path stays clean, which isolates
the cause as the GPU's FP16 precision, not the prompt, the model artifact, or
retrieval (Figure~\ref{fig:context-wall}). Forcing the GPU to FP32 also removes
the cliff, confirming it directly. The fix costs latency: FP32 prefill is
$2.1$--$2.5\times$ slower (prefill is compute-bound, and FP16 doubles Adreno
arithmetic throughput), though decode---bandwidth-bound---is essentially
unchanged, for a net $\sim$21--34\% slower total query at $k=10$--$15$.

\begin{figure}[t]
\centering
\begin{tikzpicture}[font=\footnotesize]
  % scale: 1000 tokens = 1.28cm
  \def\sx{0.00128}
  \fill[mamteal!22]   (0,0) rectangle (5000*\sx,0.62);
  \fill[mamamber!35]  (4096*\sx,0) rectangle (5000*\sx,0.62);
  \fill[red!22]       (5000*\sx,0) rectangle (6200*\sx,0.62);
  \draw[black!45] (0,0) rectangle (6200*\sx,0.62);
  % shipped cap line
  \draw[mamteal!80!black, very thick] (4096*\sx,-0.08) -- (4096*\sx,0.78);
  \draw[red!65!black, very thick] (5000*\sx,-0.08) -- (5000*\sx,0.78);
  % zone labels
  \node[text=mamteal!50!black, font=\scriptsize] at (2048*\sx,0.31) {clean (within cap)};
  \node[text=mamamber!50!black, font=\scriptsize, rotate=90] at (4548*\sx,0.31) {margin};
  \node[text=red!55!black, font=\scriptsize] at (5600*\sx,0.31) {collapse};
  % axis ticks
  \foreach \t in {0,2000,4096,5000,6000}{
    \draw[black!40] (\t*\sx,-0.08) -- (\t*\sx,-0.16);
  }
  \node[font=\scriptsize, anchor=north] at (0,-0.18) {0};
  \node[font=\scriptsize, anchor=north] at (2000*\sx,-0.18) {2k};
  \node[font=\scriptsize, anchor=north, text=mamteal!60!black] at (4096*\sx,-0.18) {\textbf{4{,}096}};
  \node[font=\scriptsize, anchor=north, text=red!60!black] at (5000*\sx,-0.18) {$\sim$5{,}000};
  \node[font=\scriptsize, anchor=north] at (6000*\sx,-0.18) {6k};
  \node[font=\scriptsize, anchor=north] at (3100*\sx,-0.52) {total context (tokens)};
  % annotations above
  \node[text=mamteal!55!black, font=\scriptsize, anchor=south, align=center] at (4096*\sx,0.82)
    {ship cap\\\scriptsize(longer prompts rejected)};
  \node[text=red!60!black, font=\scriptsize, anchor=south, align=center] at (5000*\sx,1.30)
    {FP16 GPU cliff\\\scriptsize(decode collapses)};
  \draw[->, red!55!black] (5000*\sx,1.28) -- (5000*\sx,0.80);
  % fp32 extends arrow
  \draw[->, black!55] (6000*\sx,0.95) -- (6700*\sx,0.95);
  \node[font=\scriptsize, text=black!55, anchor=west] at (6700*\sx,0.95) {FP32 GPU};
\end{tikzpicture}
\caption{The FP16-GPU context wall. The shipped $4{,}096$-token cap (teal) sits
$\sim$900 tokens below the $\sim$5{,}000-token cliff (red) past which FP16 GPU
decode deterministically collapses; the margin (amber) is the safety buffer.
The CPU FP32 path has no cliff, and forcing the GPU to FP32 extends the clean
zone at $\sim$25\% added latency.}
\label{fig:context-wall}
\end{figure}

The $4{,}096$ cap therefore buys a $\sim$900-token safety margin below the
cliff on the fast FP16 path. It is also why $k=20$ is absent from
Figure~\ref{fig:latency-k}: at that depth the 8 longest of 18 query types
produce prompts exceeding $4{,}096$ tokens and the runtime rejects them outright
(a fixed 24 of 54 runs in every $k=20$ cell), so a $k=20$ median would be
computed only over the queries short enough to survive---a survivorship
artifact rather than a real measurement. For deployments that need longer
context, the FP32 GPU path is the clean fallback at modest cost; on the FP16 GPU
path, $4{,}096$ tokens is the safe ceiling. The cliff is specific to FP16: the
shipped CPU default decodes in FP32 and never hits it, so the cap is a uniform
runtime setting that chiefly protects the GPU opt-in.

\subsection{Device compatibility}
\label{sec:devices}

Latency is comfortable on a flagship; the harder question is what a device must
provide to run the system at all. The binding constraints are not speed but
memory, storage, and GPU availability---all properties we can state directly
from the artifacts.

\paragraph{Memory and storage.} The generator's runtime memory is the hard
floor: $\sim$3.3\,GB for E4B and $\sim$1.7\,GB for E2B (the published model
specs). Android and the app add $\sim$1.5--2\,GB on top, which sets the device
RAM floor. Storage is the one-time install: the $\sim$193\,MB app, the model
weights ($3.66$\,GB for E4B, $2.59$\,GB for E2B), and the $\sim$260\,MB
knowledge base (the embedder plus the $63{,}650$-passage index)---$\sim$4.1\,GB
of assets for E4B and $\sim$3.0\,GB for E2B. These assets are obtained
once---downloaded over Wi-Fi or \emph{sideloaded} onto the device directly (USB
or memory card, needing no connectivity)---after which the system runs fully
offline; sideloading lets the install reach a low-connectivity site without
each device pulling several gigabytes over the network. Table~\ref{tab:devices}
gives the minimum and recommended specifications that follow.

\begin{table}[t]
\caption{Minimum and recommended device specifications, derived directly from
the model runtime footprints (plus Android/app overhead) and the install size
(plus working headroom). The minimum is what runs; the recommended figure
leaves headroom against OOM kills under multitasking and against future bundle
growth. These are device \emph{requirements}; we make no claim about which
retail devices meet them.}
\label{tab:devices}
\centering
\small
\setlength{\tabcolsep}{6pt}
\begin{tabular}{@{}lcccc@{}}
\toprule
 & \multicolumn{2}{c}{Device RAM} & \multicolumn{2}{c}{Free storage} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
Generator & Minimum & Recommended & Minimum & Recommended \\
\midrule
Gemma~4 E4B & 6\,GB & 8\,GB & $\sim$4.5\,GB & 6\,GB \\
Gemma~4 E2B & 4\,GB & 6\,GB & $\sim$3.5\,GB & 5\,GB \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{GPU availability.} The $\sim$19\,s figure is the opt-in GPU path, and
assumes a working GPU---which we verified only on the Qualcomm Adreno of our
test device.
LiteRT-LM's GPU path uses OpenCL, whose support is inconsistent across GPU
families: on the ARM Mali GPUs common in non-Qualcomm SoCs it is reported to
fail silently, in which case the app falls back to CPU---a path we have not
tested (\S\ref{sec:latency-pending}). The safe planning assumption is therefore
CPU-first: at the deployed $k=3$, E4B on CPU ($\sim$43\,s) still meets the
budget, and E2B on CPU ($\sim$21\,s) covers the 4\,GB devices that cannot fit
E4B at all. The GPU's $\sim$19\,s is the best case, not a guarantee.

\subsection{What we have not measured}
\label{sec:latency-pending}

Two gaps bound these results, both disclosed rather than closed. First, every
number here comes from a \emph{single flagship} device with an Adreno GPU; the
lower-cost MediaTek hardware likely to be used in the field is untested, and its
CPU-only path will be slower than the Snapdragon~8 Elite CPU
rows above (we extrapolate $\sim$2$\times$ from published compute gaps, but have
not measured it). Second, we report steady-state latency only: per-session
battery drain and sustained-load thermal throttling---both of which matter for a
clinician using the app repeatedly through a shift---are not yet characterized.
We scope these as the first follow-up measurements (\S\ref{sec:limitations})
rather than reasons to withhold the result, since the load-bearing
conclusion---fully offline answers within the field budget---holds on the
hardware we did measure.
