\section{Discussion}
\label{sec:discussion}

Read top to bottom, the evaluation inverts the intuition about where an
on-device medical RAG system is weakest. The retriever was the natural
suspect---a 300M embedder on a phone against cloud systems---yet on-device
retrieval is essentially solved (\S\ref{sec:retrieval-results}): EmbeddingGemma
ranks third of seven and the corpus holds a relevant passage for the large
majority of queries. The weak link is downstream, where it is least expected:
the small generator does not turn good retrieval into better answers. That we
can place the weak link there at all is a consequence of the evaluation design:
isolating the generator under oracle context, apart from live retrieval
(\S\ref{sec:generator}), separates a retrieval problem from a generation one,
where an end-to-end score alone would have blamed the retriever. This section
traces the finding to its causes, lays out the levers that follow, and draws out
what it means for deployment.

\subsection{The bottleneck has moved from retrieval to grounding}

The sharpest single result is also the most counter-intuitive: \emph{adding}
retrieved context does not improve answers, and the more capable the generator,
the more it hurts---the RAG penalty deepens from $-0.01$ HealthBench
\texttt{weighted\_met} at 4B to $-0.10$ at 31B (\S\ref{sec:model-size}). It is
tempting to read this as ``the retrieved context is useless,'' but that mistakes
the probe for the question. A large model already holds the answer
parametrically, so a retrieved chunk offers it little new and some distraction;
the net is negative and grows with capability. The frontier is therefore the
\emph{wrong} test of whether the context helps, because the frontier does not
need it.

The relevant subject is the deployed small model, which has no parametric route
to specific clinical detail and so depends on retrieval for it. For that model
the net cost has at least three plausible roots, only the first of which we can
quantify:
\begin{itemize}\setlength{\itemsep}{1.5pt}
  \item it under-uses the on-target context it is given---a faithful generator
  mirrors its chunk rather than reasoning over it (\S\ref{sec:retr-corpus});
  \item the latency-capped three-chunk window may carry only a fragment of a
  multi-step protocol, which a faithful model then reproduces as a truncated
  answer; and
  \item some open-ended questions are simply not well covered by guideline
  chunks at all.
\end{itemize}
The three are coupled, and only the deployment as a whole exhibits their sum;
we do not claim a measured split.

Two facts keep this from being a counsel of despair. Offline retrieval is
genuinely good---the corpus holds a relevant chunk for $96.5\%$ of queries and
EmbeddingGemma rivals cloud retrievers---so the raw material exists. And per
query the mechanism demonstrably works (Figure~\ref{fig:case-nausea}): the same
small model gives a specific, safe-to-act-on answer when the right chunk is
surfaced and a vague one when it is not. Whether RAG is net-negative because the
generator under-uses good context (fixable) or because the context is genuinely
low-value on this query distribution (not) is therefore open---and the evidence
leans toward the former.

That open question does not make retrieval optional. Even at a small average
cost, it is what makes an answer \emph{auditable}: every claim is tied to a
cited guideline passage a clinician can check (\S\ref{sec:corpus}), and a
faithful model grounded in a retrieved source is far less prone to confident
fabrication than one answering from parameters alone---the same property that
makes Gemma~4 safe to deploy (\S\ref{sec:generator}). Per query it is also
decisive when the corpus holds a local protocol or dose the model lacks. We
retain retrieval for safety, traceability, and local specificity; the goal of
the levers below is to make it pay in answer quality too, turning a small net
cost into a net gain.

\subsection{Three complementary levers}

The diagnosis maps onto three levers, each attacking a different root.

\paragraph{Generator grounding.} The central lever, and the one the on-device
setting most needs, is to teach the small generator to extract specific,
faithful guidance from retrieved context---to reach through retrieval what the
frontier reaches through scale, since scale is not available on the phone
(\S\ref{sec:model-size}). Concretely this is a grounding fine-tune: train the
deployed generator on (context, faithful-answer) supervision, optionally
distilled from a frontier model, so it uses on-target passages and resists
off-target ones while keeping every claim cited. The Qwen ceiling shows the
prompts and the task are already sound (\S\ref{sec:helpful-safe}); what is
missing is a small model that uses its context well. The experiment that would
settle the open question above is direct---a grounding fine-tune evaluated under
\emph{live} retrieval: if it flips RAG net-positive, under-grounding was the
cause; if a well-grounded small model still cannot beat no-RAG, the lever is
capability and corpus instead.

\paragraph{Procedure-aware retrieval and depth.} If part of the cost is the
three-chunk window splitting a protocol, the remedy is retrieval-side, and it is
really a question of how to spend a fixed token budget---the window is bounded by
latency and the context cap. Two knobs govern how much of a procedure that
budget captures: where chunk boundaries fall (around whole protocols rather than
page breaks) and how large each chunk is, both aimed at landing a complete
procedure in the passages retrieved rather than a fragment; reranking then
ensures it is the \emph{management} section that fills the budget, not a
definition or a study-exercise (the failure in
Figure~\ref{fig:case-nausea}). The budget itself is gated by hardware: the
shipped CPU default holds $k$ at three to stay within 60\,s, but the GPU opt-in
affords far more context ($k=15$ in $\sim$24\,s, \S\ref{sec:latency}), so a
faster backend---or an on-device NPU---buys procedure coverage directly. A cheap
first step is to vary depth and chunk size on the GPU path and measure whether
answer quality rises, testing the truncation hypothesis head-on. The caution is
that more retrieved text adds distraction as well as coverage, which is exactly
why this lever and generator grounding are coupled: richer context pays only for
a generator robust enough not to drown in it.

\paragraph{Query rewriting.} The cheapest lever carries no model cost, and it
acts where the system is most exposed: the retriever consumes the nurse's query
\emph{verbatim}. Real input is noisy---lay phrasing, abbreviations, partial
descriptions---and wording alone decides what is retrieved (``severe nausea''
and ``hyperemesis gravidarum'' fetch different bundles for the same patient,
\S\ref{sec:retr-corpus}). A lightweight rewriting or normalisation step that
cleans the query and maps it onto the corpus's vocabulary can surface the right
section before the generator runs, paired with brief user guidance on how to
describe a case. Because EmbeddingGemma is multilingual, the same front-end opens
\emph{cross-lingual} retrieval: rewriting or translating the query would let a
non-English (e.g.\ Swahili) question reach the corpus, and would let the corpus
draw on authoritative non-English references it cannot exploit
today---high-quality German-language obstetrics texts, for one---a concrete path
toward the multilingual support a real deployment needs
(\S\ref{sec:limitations}).

\noindent None of the three suffices alone. Corpus and query raise the precision
of what reaches the generator; grounding lets the generator use it. The corpus
itself remains a real but secondary constraint: coverage is high in aggregate,
yet it was not expanded this cycle and licensing limits some high-value
reference texts (\S\ref{sec:limitations}).

\subsection{Deployment is a safety-versus-usefulness choice}

Beyond the bottleneck, the choice of which generator to deploy carries a lesson
of its own. Between two same-size on-device candidates we did \emph{not} choose
the more capable one. Gemma~3n is the more
helpful model, but it produces genuine, confident dosing and drug errors, and
two independent probes---the targeted Kenya dangerous-answer count
(\S\ref{sec:helpful-safe}) and oracle faithfulness (\S\ref{sec:generator})---%
agree that Gemma~4 is roughly twice as safe and grounds about as faithfully as a
frontier model. So we deploy the faithful model and recover its helpfulness with
the prompt (\S\ref{sec:system-prompt}), rather than deploy the more capable
model and try to suppress its errors. In a safety-critical, low-supervision
setting, faithfulness to sources is the right \emph{primary} criterion for
choosing the generator and raw capability the secondary one---the reverse of the
usual leaderboard order.

\subsection{Open questions raised by the findings}

Beyond the levers, two design questions surface that a fielded system will have
to answer.

\paragraph{Conflicting guidance.} The corpus draws on many authoritative
sources, and they do not always agree---an issue we met concretely. Auditing the
faithfulness failures (\S\ref{sec:generator}) surfaced 16 cases in which two
retrieved passages give conflicting advice on the same clinical question, each
verified verbatim against its source. They are not edge cases: a field reference
re-doses emergency contraception after vomiting within \emph{three} hours where
WHO says \emph{two}; one WHO document starts the active phase of labour at
4\,cm dilatation where another starts it at 5\,cm; one source calls a placenta
retained after one hour where NICE uses 30/60 minutes.
Appendix~\ref{app:contradictions} reproduces the full retrieved context for
three such cases with the conflicting statements highlighted---the model is
handed both in a single prompt, so whichever it follows can be faulted against
the other. When sources diverge like this, what should the system do---prefer
the local or most recent protocol, surface both with their provenance, or flag
the conflict to the clinician? Citations make the divergence visible---the
system already shows its sources---but choosing a behaviour, and validating it
clinically, is open for any assistant grounded in a heterogeneous guideline
base.

\paragraph{Knowing when to abstain.} A deployable assistant needs a principled
way to decline, rather than guess, on a query the corpus cannot support. The
technical half is unsolved: we found no usable confidence gate today
(\S\ref{sec:retr-gate}), as neither a retrieval score nor an answer-side signal
reliably separates answerable from unanswerable queries. But the harder half is
clinical. \emph{When} the system should abstain, and \emph{what} it should do
instead, are not threshold choices: they have to be set against the
nurse-midwife's responsibility boundary---which decisions fall within their
scope of practice, and which must be escalated or referred. The expected
behaviour at that boundary---decline, hand off with a safe holding action, or
surface the uncertainty---must be defined carefully and with clinicians before
any gate is tuned to enforce it.

These sit alongside the prerequisites for any real deployment---Swahili support,
validation on the lower-cost hardware of the field, and above all a study with
nurse-midwives in place of LLM-judge proxies---which we treat as limitations
(\S\ref{sec:limitations}) rather than open research questions.

