\section{Limitations}
\label{sec:limitations}

MAM-AI is a thoroughly evaluated research prototype, not a fielded product: all
our evidence is benchmark behaviour scored by LLM judges, and several gaps bound
what that evidence can claim.

\begin{itemize}\setlength{\itemsep}{3pt}
  \item \textbf{No field test, and no real users.} A lag in project funding
  meant the planned study with Zanzibar nurse-midwives could not proceed this
  cycle, so there is no point-of-care data on how the system performs in real
  clinical use.

  \item \textbf{Not clinician-verified.} Clinical correctness is not
  established by humans. Every quality, safety, and faithfulness number comes
  from LLM judges; the rubric judge's agreement with physicians was measured
  only on a third-party benchmark's physician-labelled data (HealthBench,
  \S\ref{sec:judge}), not on clinicians reviewing MAM-AI's own answers. The
  dangerous-answer determinations are likewise model-generated, awaiting review
  by qualified clinicians.

  \item \textbf{English only.} The system, knowledge base, and evaluation are
  entirely in English, while the deployment setting is Swahili-speaking and the
  bilingual prompt was withdrawn as unvalidated (\S\ref{sec:system-prompt});
  Swahili support remains an open deployment gap.

  \item \textbf{Retrieval does not yet improve answers.} On the small on-device
  generator, better retrieval does not reach the answers and added context is
  net-neutral-to-negative (\S\ref{sec:retrieval-results}); we retain retrieval
  for traceability and per-query specificity, not for average quality, and
  closing this gap---generator-side grounding---is unsolved
  (\S\ref{sec:future}).

  \item \textbf{Latency measured on a single device.} All timings come from one
  flagship (Snapdragon~8 Elite); the lower-cost MediaTek hardware likely used in
  the field is untested and will be slower---possibly past budget on the shipped
  CPU path---and battery drain and sustained thermal throttling on a warm ward
  are not yet profiled (\S\ref{sec:latency-pending}).

  \item \textbf{Corpus coverage and quality.} The corpus was not expanded this
  cycle, so coverage gaps remain; it contains internal contradictions between
  sources (\S\ref{sec:future}); and licensing limits some high-value
  reference texts.
\end{itemize}
