\section{Related Work}
\label{sec:related}

\paragraph{Clinical AI in low-resource settings.} The closest precedents to
MAM-AI are point-of-care digital tools for frontline maternal-health workers.
The Safe Delivery App is an \emph{offline} smartphone reference for birth
attendants---curated text, animations, and instructional video---that more than
doubled postpartum-haemorrhage management skills in a cluster-randomised trial in
Ethiopia~\cite{christiansen2023safedelivery}; it shares MAM-AI's offline,
point-of-care setting, but is a fixed-content library rather than an assistant
that answers free-form questions. More recently, LLM clinical decision support
has been deployed in African primary care, reducing diagnostic and treatment
errors across tens of thousands of visits in Kenya~\cite{korom2025clinical},
though cloud-served and clinician-facing in facilities. MAM-AI joins these two
threads---an offline, on-device LLM that answers nurse-midwives'
questions---the regime closest to MedAide, an offline edge-LLM medical
assistant~\cite{basit2024medaide}. Pan-African medical QA benchmarks such as
AfriMed-QA~\cite{olatunji2025afrimedqa} motivate evaluation for these settings,
but are datasets rather than deployed systems.

\paragraph{Medical language models and retrieval.} Open and frontier medical
LLMs---among them the Gemma-based MedGemma~\cite{sellergren2025medgemma}---are
large and server-hosted; we instead target a 4B model that runs on-device, and
find (\S\ref{sec:model-size}) that capability dominates quality while
faithfulness decides which model we deploy. Retrieval-augmented clinical systems
such as Almanac~\cite{zakka2024almanac} argue that grounding answers in cited
sources is what makes a medical assistant trustworthy---the same motivation
behind our traceable RAG---and medical-RAG benchmarks~\cite{xiong2024benchmarking}
dissect retriever and generator contributions; unlike that cloud setting, we
find the small \emph{generator}, not retrieval, to be the bottleneck. We
evaluate on HealthBench~\cite{arora2025healthbench}, whose physician-written
rubrics let us separate completeness from safety.

\paragraph{On-device inference and quantization.} Running LLMs on commodity
mobile devices is enabled by mobile runtimes---Google's LiteRT-LM~\cite{litertlm2025}
(our generator's runtime), and open engines such as llama.cpp and
MLC-LLM---together with low-bit quantization: weight-only int4
methods~\cite{lin2024awq} and the outlier-aware 8-bit analysis
of~\citet{dettmers2022llmint8}, whose findings on low-precision transformer
numerics bear directly on the FP16 attention collapse we characterise
(\S\ref{sec:context-wall}). Where much on-device work favours sub-billion-%
parameter models, we deploy a 4B model at int4, trading precision for capability.

\paragraph{Faithfulness and judge evaluation.} Detecting whether a generation
stays faithful to its source spans claim-level fact-checking~\cite{tang2024minicheck}
and RAG-specific faithfulness metrics~\cite{es2024ragas}; we use Patronus
Lynx~\cite{ravi2024lynx} as the first stage of our two-pass faithfulness
pipeline. Our rubric and safety scores come from an LLM-as-judge, a standard but
bias-prone practice~\cite{zheng2023judging}, so we validate the judge against
physician labels (\S\ref{sec:judge}) rather than trust it blind.

\paragraph{On-device text embeddings.} Distillation has made small embedders
competitive with much larger ones---Gecko~\cite{lee2024gecko} and the 300M
EmbeddingGemma~\cite{schechtervera2025embeddinggemma}, our retriever---and our
results add evidence that an on-device embedder rivals cloud APIs on medical
retrieval. We benchmark against general (MTEB~\cite{muennighoff2023mteb}) and
biomedical (MedCPT~\cite{jin2023medcpt}) baselines, but with graded relevance and
Precision@$k$ tailored to the deployment task.
