\section{Future Directions}
\label{sec:future}

The evaluation points to a concrete programme of work, in three tracks:
near-term technical levers that follow directly from the diagnosis
(\S\ref{sec:discussion}); open design questions that need clinical partners as
much as engineering; and the prerequisites for a real field deployment. We order
the levers \emph{foundation-first}, working from the corpus that everything else
depends on out to the generator that consumes it.

\subsection{Technical levers}

We take them from the foundation outward---the corpus everything retrieves from,
the queries that hit it, the languages it can reach, and the generator's use of
what it receives---then close with two retrieval-tuning knobs, chunking and
depth.

\paragraph{Corpus expansion and curation (the foundation).} A RAG system is only
as good as the guidelines it can retrieve, and a small generator that leans on
what it receives (\S\ref{sec:retrieval-results}) makes the corpus decisive.
Coverage is high in aggregate---a relevant chunk exists for $96.5\%$ of
queries---but the corpus was not expanded this cycle, and per-query gaps are
decisive when they fall on a query it cannot answer. The reliable, direct lever
is to grow the corpus on two fronts---authoritative international guidelines
(WHO, NICE, and comparable bodies) and local protocols (Tanzania and
Zanzibar)---\emph{broadening} each so more questions have a trustworthy source at
all and \emph{deepening} each toward information-rich, actionable content: the
exact doses, thresholds, and steps a nurse can act on, including the locally
specific ones. It is the slowest lever but the most reliable, and its payoff is the
most direct per query.

\paragraph{Query rewriting.} The retriever consumes the nurse's query
\emph{verbatim}, and real input is noisy---lay phrasing, abbreviations, partial
descriptions---where wording alone decides what is retrieved (``severe nausea''
and ``hyperemesis gravidarum'' fetch different bundles for the same patient,
\S\ref{sec:retr-corpus}). A rewriting or normalisation step that maps the query
onto the corpus's vocabulary can surface the right section before the generator
runs, paired with brief user guidance on how to describe a case. Its cost is an
extra inference step on the device: cheap to build but not free at run time---it
adds latency (and, as a separate model, memory) to an already-tight budget---so
it must earn its place through a retrieval-precision gain, and is most attractive
when it reuses the deployed model to emit only a short normalised query rather
than loading a second model.

\paragraph{Multilingual and cross-lingual retrieval.} Because EmbeddingGemma and
Gemma~4 are both multilingual, the system can in principle retrieve and answer
across languages: an English query fetching German guidelines, or a Swahili query
fetching English and German passages and being answered in Swahili. This is
doubly valuable---it would let the corpus draw on authoritative non-English
references it cannot exploit today (high-quality German-language obstetrics
texts, for one), feeding the corpus lever above, and it is the natural path
toward serving Swahili-speaking users. It is also the most \emph{exploratory}
lever here: cross-lingual retrieval quality and the faithfulness of Swahili
answers are unmeasured and would have to be evaluated before being relied on.
(This is a research lever, distinct from \emph{shipping} a validated Swahili
configuration: the latter is what field use requires, and we list it separately
among the deployment prerequisites below.)

\paragraph{Generator grounding (the deepest lever).} Once the corpus is rich and
the right passages reach the model, the remaining lever is the generator's use of
them: teach the small model to extract specific, faithful guidance from retrieved
context---to reach through retrieval what the frontier reaches through scale,
since scale is not available on the device (\S\ref{sec:model-size}). Concretely
this is a grounding fine-tune: train the deployed generator on
(context, faithful-answer) supervision, optionally distilled from a frontier
model, so it uses on-target passages and resists off-target ones while keeping
every claim cited. Unlike the retrieval-side levers it adds no run-time
cost---the same model, one pass, better weights---but it is the most expensive to
build and depends on the corpus work above for its training data. The Qwen
ceiling shows the prompts and task are already sound (\S\ref{sec:helpful-safe});
what is missing is a small model that uses its context well. The experiment that
would settle whether RAG can be made net-positive is direct---a grounding
fine-tune evaluated under \emph{live} retrieval: if it flips RAG net-positive,
under-grounding was the cause; if a well-grounded small model still cannot beat
no-RAG, the remaining constraint is capability and corpus.

\paragraph{Chunking and retrieval depth.} Two retrieval-shaping knobs tune how
well the retrieved content serves a query, and---unlike the levers above---they
pay off \emph{only} once the corpus holds good guidelines and only within the
latency budget. \emph{Chunking} that keeps a whole protocol intact, rather than
splitting it across chunks, avoids handing the generator a fragment it can answer
only partially; greater \emph{retrieval depth} (a larger $k$) brings more of a
procedure into the window but spends latency---the shipped CPU path holds $k$ at
three for the 60\,s budget, while a GPU affords more ($k=15$ in
$\sim$24\,s, \S\ref{sec:latency}). Reranking complements both, ensuring the
\emph{management} section fills the budget, not a definition or a study-exercise
(Figure~\ref{fig:case-nausea}). These are tuning levers, not foundations---listed
last because they are gated on the corpus and the hardware above them.

\subsection{Open design questions}

Two design questions a fielded system must answer need clinical partners as much
as engineering.

\paragraph{Knowing when to abstain.} Sometimes the safest answer is no answer:
when the corpus cannot support a query, the system should say so rather than
guess. Building this has two parts. The first is technical---spotting the queries
the system is likely to get wrong---and it is unsolved: none of the signals we
tried tell answerable queries from unanswerable ones, neither a retrieval score
nor an answer-side check (\S\ref{sec:retr-gate}). A usable signal would probably
have to come from the generator itself---how confident or self-consistent its
answer is---which we have not built. The second part is clinical, and harder:
deciding \emph{when} the system should hold back, and \emph{what} it should tell
the nurse instead, depends on the nurse-midwife's scope of practice---which
decisions are theirs to make, and which must be referred to a doctor. That line
has to be drawn with clinicians before any such gate is built.

\paragraph{Conflicting guidance.} The corpus draws on many authoritative
sources, and they do not always agree. Auditing the faithfulness failures
(\S\ref{sec:generator}) surfaced 16 cases in which two retrieved passages give
conflicting advice on the same clinical question, each verified verbatim against
its source---a field reference re-doses emergency contraception after vomiting
within \emph{three} hours where WHO says \emph{two}; one WHO document starts the
active phase of labour at 4\,cm dilatation where another starts it at 5\,cm
(Appendix~\ref{app:contradictions}). These are few, and whether the differences
matter clinically is not ours to judge---that needs clinicians. The mechanism,
though, is real: the model is handed both passages at once, so whichever it
follows can be faulted against the other. When sources diverge,
what should the system do---prefer the local or most recent protocol, surface
both with their provenance, or flag the conflict to the clinician? Citations make
the disagreement visible, but deciding what the system should do about it---and
confirming that choice with clinicians---is still unsolved, and not just for
MAM-AI: any assistant built on many different sources faces it.

\subsection{Toward a field deployment}

Beyond the research levers, a real deployment has prerequisites this cycle could
not close; they bound what our evidence can claim (\S\ref{sec:limitations}) and
form the engineering-and-study track a successor would run in parallel.

\begin{itemize}\setlength{\itemsep}{2.5pt}
  \item \textbf{A field study with nurse-midwives.} Above all, the system needs
  a study with the nurse-midwives it is built for: all current evidence is
  benchmark behaviour scored by LLM judges, with no point-of-care data on real
  clinical use. This is the single most important gap.
  \item \textbf{Clinician verification.} Every quality, safety, and faithfulness
  number comes from LLM judges; the dangerous-answer determinations in particular
  await review by qualified clinicians. Independent clinical verification is a
  prerequisite for any claim of clinical correctness.
  \item \textbf{Swahili support.} The system, corpus, and evaluation are
  English-only, while the deployment setting is Swahili-speaking, and a
  safety-critical prompt cannot ship in a language it was not evaluated in
  (\S\ref{sec:system-prompt}). Validating a Swahili configuration---which the
  cross-lingual lever above complements---is a prerequisite for field use.
  \item \textbf{Validation on field hardware.} All timings come from one flagship
  device---a OnePlus tablet with a Snapdragon~8 Elite SoC; the lower-cost
  MediaTek devices likely used in the field are untested and will be slower, and
  battery drain and sustained thermal throttling over a ward shift are unprofiled
  (\S\ref{sec:latency-pending}). A smaller Gemma~4 E2B variant is available as a
  fallback for memory-constrained or slower devices---it roughly halves the
  runtime memory footprint (dropping the device RAM floor from 6 to 4\,GB) and is
  somewhat faster (up to ${\sim}2\times$ on CPU total)---but it too has been
  measured only on this flagship. We also timed the baseline prompt, whereas the
  release uses the G1 prompt, which can produce longer responses and thus more
  latency, so the measured numbers are a lower bound.
\end{itemize}
